### 2021-05-31 manuscript/main.tex
added the abstract

\documentclass{bioinfo}

\usepackage{relsize} % for Cpp
\usepackage{xspace} % for Cpp

\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

\copyrightyear{2021} \pubyear{2021}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Original Paper}

\begin{document}
    \firstpage{1}

    \subtitle{Sequence analysis}

%\title[short Title]{Wavefront inception scales global pairwise alignment to whole eukaryotic chromosomes}
%\title[short Title]{Wavefront inception enables global pairwise alignment to whole eukaryotic chromosomes}
    \title[wfmash: whole-cromosome pairwise alignment]{Whole-chromosome pairwise alignment using a hierarchical wavefront algorithm}
    \author[Sample \textit{et~al}.]{
        Corresponding Author\,$^{\text{\sfb 1,}*}$,
        Co-Author\,$^{\text{\sfb 2}}$ and
        Co-Author\,$^{\text{\sfb 3,}*}$
    }
    \address{
        $^{\text{\sf 1}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 3}}$Department, Institution, City, Post Code, Country
    }

    \corresp{$^\ast$To whom correspondence should be addressed.}

    \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

    \editor{Associate Editor: XXXXXXX}

    \abstract{
        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
        including pangenomes building. Pangenomes are sequence models able to provide a full representation of the
        mutual alignment of collections of genomes. The time and memory required to compute a pairwise alignment
        increase quadratically with the sequence length, making it impractical to directly align very long sequences
        without applying heuristic approaches to determine possible syntenic regions in the sequences. Nevertheless,
        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
        the development of tools able to align sequences of the order of tens of megabases long.
        \\
        \textbf{Results:} Here we present wfmash, a new gap-affine pairwise aligner designed to align DNA sequences
        at a whole-chromosome scale. wfmash applies a hierarchical implementation of the wavefront alignment
        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
        time and memory.
        \\
        \textbf{Availability:} wfmash is written in \Cpp\ programming language. Source code and user manual are freely available at
        \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
        \\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

    \maketitle

    \section{Introduction}

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text. Figure~\ref{fig:01}
    shows that the above method Text Text Text Text Text Text Text
    Text Text Text Text Text.  \citep{Bag01} wants to know about
        {\ldots}{\ldots} text follows.
    \begin{equation}
        \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
    \end{equation}

%\enlargethispage{12pt}

    \section{Approach}

    Equation~(\ref{eq:01}) Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    .....

    \begin{methods}
        \section{Methods}

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text
        text\vspace*{1pt}

        \begin{itemize}
            \item for bulleted list, use itemize
            \item for bulleted list, use itemize
            \item for bulleted list, use itemize\vspace*{1pt}
        \end{itemize}

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text\vadjust{\newpage}.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.

        \subsection{This is subheading}

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text
        text. Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.

        \subsubsection{This is subsubheading}

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text\break
        text text

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text
        textText Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.

        \enlargethispage{6pt}

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.

        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text
        Text Text Text Text Text Text Text Text Text Text Text Text Text
        Text Text Text Text Text Text Text Text.
        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
        Text Text Text Text Text Text Text Text Text Text Text.
        \citealp{Boffelli03} might want to know about text text text text

        \begin{table}[!t]
            \processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}
                                                                    \toprule head1 &
                                                                    head2 & head3 & head4 \\\midrule
                                                                    row1 & row1 & row1 & row1 \\
                                                                    row2 & row2 & row2 & row2 \\
                                                                    row3 & row3 & row3 & row3 \\
                                                                    row4 & row4 & row4 & row4 \\\botrule
            \end{tabular}}{This is a footnote}
        \end{table}

    \end{methods}

    \begin{figure}[!tpb]%figure1
        \fboxsep=0pt\colorbox{gray}{\begin{minipage}[t]{235pt}
                                        \vbox to 100pt{\vfill\hbox to
                                        235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}\vfill}
        \end{minipage}}
%\centerline{\includegraphics{fig01.eps}}
        \caption{Caption, caption.}\label{fig:01}
    \end{figure}

%\begin{figure}[!tpb]%figure2
%%\centerline{\includegraphics{fig02.eps}}
%\caption{Caption, caption.}\label{fig:02}
%\end{figure}

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text

    \subsection{Test1}

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text

    \section{Discussion}

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text.

    Table~\ref{Tab:01} shows that Text Text Text Text Text Text Text
    Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
    the above method Text Text. Text Text Text Text Text Text Text
    Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
    method Text Text. Text Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \section{Conclusion}

    (Table~\ref{Tab:01}) Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text

    \begin{enumerate}
        \item this is item, use enumerate
        \item this is item, use enumerate
        \item this is item, use enumerate
    \end{enumerate}

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
    above method Text Text Text Text Text Text Text Text Text Text
    Text Text.  \citealp{Boffelli03} might want to know about text
    text text text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text Text Text Text Text Text Text Text Text.
    \citealp{Boffelli03} might want to know about text text text text
    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text\break Text.

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text.
    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
    Text Text Text\vspace*{-10pt}

    \section*{Acknowledgements}

    Text Text Text Text Text Text Text Text.  \citealp{Boffelli03} might want to know about text
    text text text\vspace*{-12pt}

    \section*{Funding}

    This work has been supported by the... Text Text Text Text.\vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

    \begin{thebibliography}{}

        \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
        Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

        \bibitem[Bag {\it et~al}., 2001]{Bag01}
        Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

        \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
        Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
        in nigral dopaminergic neurnol cell: correlation with the known
        pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
        Res.}, \textbf{110}(Suppl. 1), 76--84.

        \bibitem[Lehmann, 1986]{Leh86}
        Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

        \bibitem[Crenshaw and Jones, 2003]{Cre03}
        Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
        cancer management: one tumor, one chip. \textit{Bioinformatics},
        doi:10.1093/bioinformatics/btn000.

        \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
        Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
        (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
        ???--???.

        \bibitem[Bardet, 1920]{Bar20}
        Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
        polydactylie et retinite pigmentaire (contribution a l'etude des
        formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
        institution, Paris, France.

    \end{thebibliography}
\end{document}

### 2021-05-31 manuscript/main.tex
added the introduction

@@
 
+\newcommand{\wfmash}{wfmash}
 \newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
@@
         \\
-        \textbf{Results:} Here we present wfmash, a new gap-affine pairwise aligner designed to align DNA sequences
-        at a whole-chromosome scale. wfmash applies a hierarchical implementation of the wavefront alignment
+        \textbf{Results:} Here we present \wfmash\, a new gap-affine pairwise aligner designed to align DNA sequences
+        at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
         algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
@@
         \\
-        \textbf{Availability:} wfmash is written in \Cpp\ programming language. Source code and user manual are freely available at
+        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely available at
         \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
@@
 
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text. Figure~\ref{fig:01}
-    shows that the above method Text Text Text Text Text Text Text
-    Text Text Text Text Text.  \citep{Bag01} wants to know about
-        {\ldots}{\ldots} text follows.
-    \begin{equation}
-        \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
-    \end{equation}
+    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping
+    \citep{22388286, BWA_MEM, 23103880} to variant detection~\citep{21478889}, as well as de~novo
+    assembly \citep{19251739} and pangenome building~\citep{33177663, 33066802}. Pairwise sequence alignment
+    can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary
+    relationships between two biological sequences. In this context, pangenomes model the full set
+    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
+    relationships between all the genomes represented, in contrast to reference-based approaches which
+    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
+    the entire genetic diversity of a population. This opportunity, in combination with the massive amount
+    of data available, made possible thanks to the advancement in sequencing technology, presses for the
+    development of tools able to compute pairwise alignment of very long sequences.
+
+    To face the upcoming challenges in pangenomes building, we have developed \wfmash\, a new gap-affine
+    pairwise tool for aligning DNA sequences at a whole-chromosome scale. It supports split-read alignment
+    and gap-affine penalties for insertion and deletions. \wfmash\ scales efficiently to large eukaryotic
+    chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise
+    alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the integration of
+    MashMap~\citep{30423094}, a fast approximate aligner for computing local alignment boundaries between
+    long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine algorithm
+    that takes advantage of homologous regions between the sequences to accelerate the alignment process.
+
+    %TODO Some sentences of why it exists with respect to Minimap2/Winnomap2, explaining the limitations of those tools in the pairwise alignment task.
 

### 2021-05-31 manuscript/main.tex
added the methods; cleaning

@@
 
-    \section{Approach}
-
-    Equation~(\ref{eq:01}) Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    .....
-
-    \begin{methods}
     \section{Methods}
 
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text
-        text\vspace*{1pt}
+    \wfmash\ first applies a locality-sensitive hashing, from MashMap, to rapidly determine syntenic region
+    boundaries between long DNA sequences. Then, a hierarchical implementation of the WFA allows computing the
+    base-level global alignment of the identified mappings.
 
-        \begin{itemize}
-            \item for bulleted list, use itemize
-            \item for bulleted list, use itemize
-            \item for bulleted list, use itemize\vspace*{1pt}
-        \end{itemize}
+    \subsection{Approximate mapping}
 
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
+    Each query sequence is broken into non-overlapping pieces of the requested length. These segments are then
+    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, fixing bugs
+    and incorporating the robust winnowing \citep{Schleimer S. et al.  (2003)} in the minimizer downsampling.
+    Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
+    runtime and memory-usage without affecting accuracy~\citep{32657365}.
 
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text\vadjust{\newpage}.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
+    %TODO THINGS ON THE FILTERING?
+    %TODO THING ON THE SPLIT READ AND/OR THE MERGING?
 
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
+    In addition to the segment length, the minimum segment estimated identity and the maximum number of
+    mappings to report for each segment can be specified in the mapping process. These settings allow users
+    to precisely define the mapping space to consider, specifying the characteristics of homologies to compute.
+    Each mapping location is then used as a target for base-level alignment using a hierarchical implementation
+    of the WFA.
 
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
+    \subsection{Base-level alignment}
 
-        \subsection{This is subheading}
+    The time and memory required to compute the base-level alignment increases quadratically with the sequence
+    length. The WFA provides an efficient way to decrease the amount of computation required to obtain the
+    optimal alignment between two sequences, reducing the cost to be quadratic in the alignment penalty score
+    of the optimal alignment~\citep{32915952}. This means that the algorithm is very efficient in aligning
+    similar homologous sequences (i.e., sequences with a low alignment penalty score), but high divergence
+    genomes and/or noisy long reads can increase its memory and runtime costs.
 
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text
-        text. Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-
-        \subsubsection{This is subsubheading}
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text\break
-        text text
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text
-        textText Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-
-        \enlargethispage{6pt}
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-        Text Text Text Text Text Text Text Text Text Text Text Text Text
-        Text Text Text Text Text Text Text Text.
-        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-        Text Text Text Text Text Text Text Text Text Text Text.
-        \citealp{Boffelli03} might want to know about text text text text
-
-        \begin{table}[!t]
-            \processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}
-                                                                    \toprule head1 &
-                                                                    head2 & head3 & head4 \\\midrule
-                                                                    row1 & row1 & row1 & row1 \\
-                                                                    row2 & row2 & row2 & row2 \\
-                                                                    row3 & row3 & row3 & row3 \\
-                                                                    row4 & row4 & row4 & row4 \\\botrule
-            \end{tabular}}{This is a footnote}
-        \end{table}
-
-    \end{methods}
+    To avoid such limitation, wfmash applies a hierarchical WFA, exploiting the WFA to guide the alignment
+    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
+    sequences, it aligns them to each other in small pieces W-bps long by applying a global WFA alignment
+    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
+    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
+    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
+    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
+    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
+    two fragments.
 
     \begin{figure}[!tpb]%figure1
-        \fboxsep=0pt\colorbox{gray}{\begin{minipage}[t]{235pt}
+        \fboxsep=0pt\colorbox{gray}{
+            \begin{minipage}[t]{235pt}
                 \vbox to 100pt{\vfill\hbox to
@@
             %\centerline{\includegraphics{fig01.eps}}
-        \caption{Caption, caption.}\label{fig:01}
+        \caption{Generate a cool plot using R as a base to create a cool image.}\label{fig:1}
     \end{figure}
@@
 
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
+    Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments
+    on the optimal path in the high order DP-matrix. During the process, a further step is performed for
+    refining the alignment. Indeed, inaccurate mapping estimation can lead to under-alignment at the beginning
+    and the end of the aligned syntenic region. Therefore, edlib is applied by computing semi-global alignments
+    to resolve the mapping boundaries.
+    The hierarchical implementation requires only the memory to align a syntenic region at \textit{W}-bps resolution,
+    limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
+    This approach is more flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
+    no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
+    %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
+    %TODO 'and can benefit from parallel exploration of the wavefront.' Can we?
 
-    \subsection{Test1}
+    \section{Results}
 
@@
     Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-
-    \section{Discussion}
-
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text.
-
-    Table~\ref{Tab:01} shows that Text Text Text Text Text Text Text
-    Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
-    the above method Text Text. Text Text Text Text Text Text Text
-    Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
-    method Text Text. Text Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text.
 
@@
 
-    \section{Conclusion}
-
-    (Table~\ref{Tab:01}) Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
+    \section{Discussion}
 
@@
     Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-
-    \begin{enumerate}
-        \item this is item, use enumerate
-        \item this is item, use enumerate
-        \item this is item, use enumerate
-    \end{enumerate}
-
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
-    above method Text Text Text Text Text Text Text Text Text Text
-    Text Text.  \citealp{Boffelli03} might want to know about text
-    text text text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text Text Text Text Text Text Text Text Text.
-    \citealp{Boffelli03} might want to know about text text text text
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text\break Text.
-
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-    Text Text Text\vspace*{-10pt}
 

### 2021-05-31 manuscript/main.tex
added the discussion

@@
 
+    \section{Discussion}
+
+    We implemented a novel gap-affine pairwise aligner, \wfmash\, to accelerate the computation of the mutual
+    alignment of collections of genomes, a step required for building pangenome models. We have demonstrated that
+    it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
+    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, wfmash is already
+    successfully applied in our pangenome graphs building pipeline (\citep{pggb}). No less important, pairwise
+    alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the
+    increasing yields of sequencing technologies in the coming years.
+
     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    \section{Discussion}
-
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
-
-    \section*{Acknowledgements}
-
-    Text Text Text Text Text Text Text Text.  \citealp{Boffelli03} might want to know about text
-    text text text\vspace*{-12pt}
-
-    \section*{Funding}
-
-    This work has been supported by the... Text Text Text Text.\vspace*{-12pt}
 

### 2021-05-31 manuscript/main.tex
added Data availability

@@
 
+    \vspace*{+12pt}
+    \textit{Financial Support}: none declared.
+
+    \textit{Conflict of Interest}: none declared.
+    \vspace*{+12pt}
+
+    \section*{Data availability}
+
+    Data used for the simulations is available at \href{https://github.com/marbl/CHM13}{https://github.com/marbl/CHM13}.
+    Data used for Human pangenome graph analysis is available at \href{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}{https://github.com/human-pangenomics/HPP\_Year1\_Data\_Freeze\_v1.0}.
+
+    \section*{Acknowledgements}
+    %ToDo
+    XXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
+
+    \section*{Funding}
+    %ToDo
+    This work has been supported by the XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
+
 %\bibliographystyle{natbib}

### 2021-06-01 manuscript/main.tex
started the Result section

@@
 
-    Text Text Text Text Text Text Text Text Text Text Text Text Text
-    Text Text Text Text Text Text Text Text.
+    We evaluated the performance of \wfmash\ 0.xx against Minimap2-v2.20-r1061 {29750242} and Winnomap 2.03 {XXX}, in
+    terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
+    data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
+    The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K
+    CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.
+
+    \subsection{Simulated data}
+    We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and
+    the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of divergence and length
+    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool~\citep{splitfa}. In each
+    run, all sequences have the same length. Single nucleotide variants (SNVs), small indels, and structural
+    variants (SVs) were introduced using the Mutation-Simulator script~\citep{32780803}.
+    %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.
+
+    To evaluate the correctness of the alignments, we used the pairwise alignment of the input sequences to build
+    a pangenome graph in each run, and called the variants on the graphs. The graphs were built using~\citep{seqwish}.
+    seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
+    sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
+    evaluating the false-negative and false-positive rates (FNR, FPR) using vcfeval~\citep{vcfeval}.
+
+    \begin{table}[!t]
+        \processtable{
+            \small
+            Performance of pairwise alignment of long sequences from the \textit{Saccharomyces cerevisiae}
+            chromosome IV.
+            \label{Tab:01}} {
+            \begin{tabular}{@{}llllllll@{}}
+                \toprule Aligner & Divergence & Length & Runtime (mm:ss) & Memory (GB) & Precision & Sensitivity & F-measure \\
+                \midrule
+                row1             & row1       & row1   & row1            & row1        & row1      & row1        & row1      \\
+                row2             & row2       & row2   & row2            & row1        & row1      & row1        & row1      \\
+                row3             & row3       & row3   & row3            & row1        & row1      & row1        & row1      \\
+                row4             & row4       & row4   & row4            & row1        & row1      & row1        & row1      \\
+                \botrule
+            \end{tabular}
+        }
+
+    \end{table}
+
+    \subsection{Real data}
+    % HPRC data time/memory
+    % Histogram of the contig length distribution (supplementary?)
 

### 2021-06-12 manuscript/main.tex
little fixes; notes

@@
         \\
-        \textbf{Results:} Here we present \wfmash\, a new gap-affine pairwise aligner designed to align DNA sequences
+        \textbf{Results:} Here we present \wfmash\/, a new gap-affine pairwise aligner designed to align DNA sequences
         at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
@@
     relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
-    the entire genetic diversity of a population. This opportunity, in combination with the massive amount
-    of data available, made possible thanks to the advancement in sequencing technology, presses for the
+    the entire genetic diversity of a population. FIXME In combination with the massive amount
+    of data available made possible thanks to the advancement in sequencing technology, presses for the
     development of tools able to compute pairwise alignment of very long sequences.
 
-    To face the upcoming challenges in pangenomes building, we have developed \wfmash\, a new gap-affine
+    To face the upcoming challenges in pangenomes building, we have developed \wfmash\/, a new gap-affine
     pairwise tool for aligning DNA sequences at a whole-chromosome scale. It supports split-read alignment
@@
     Each query sequence is broken into non-overlapping pieces of the requested length. These segments are then
-    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, fixing bugs
-    and incorporating the robust winnowing \citep{Schleimer S. et al.  (2003)} in the minimizer downsampling.
+    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, incorporating
+    the robust winnowing~\citep{Schleimer S. et al. (2003)} in the minimizer downsampling.
     Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
@@
     process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
-    sequences, it aligns them to each other in small pieces W-bps long by applying a global WFA alignment
+    sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment
     (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
@@
     to resolve the mapping boundaries.
+
     The hierarchical implementation requires only the memory to align a syntenic region at \textit{W}-bps resolution,
     limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
-    This approach is more flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
-    no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
+    This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
+    no heuristic seeding step, and FIXME can benefit from parallel exploration of the wavefront.
     %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
@@
 
-    We evaluated the performance of \wfmash\ 0.xx against Minimap2-v2.20-r1061 {29750242} and Winnomap 2.03 {XXX}, in
+    We evaluated the performance of \wfmash\ 0.xx against Minimap2-v2.20-r1061 \citep{29750242} and Winnomap 2.03 \citep{XXX}, in
     terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
@@
     the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of divergence and length
-    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool~\citep{splitfa}. In each
-    run, all sequences have the same length. Single nucleotide variants (SNVs), small indels, and structural
-    variants (SVs) were introduced using the Mutation-Simulator script~\citep{32780803}.
+    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool~\citep{splitfa},
+    In each run, all sequences have the same length, with 50\% of them in reverse complement.
+    Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced using the
+    Mutation-Simulator script~\citep{32780803}.
     %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.
@@
     To evaluate the correctness of the alignments, we used the pairwise alignment of the input sequences to build
-    a pangenome graph in each run, and called the variants on the graphs. The graphs were built using~\citep{seqwish}.
+    a pangenome graph in each run, identifying the variants embedded in the graphs. The graphs were built using~\citep{seqwish}.
     seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
     sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
-    evaluating the false-negative and false-positive rates (FNR, FPR) using vcfeval~\citep{vcfeval}.
+    evaluating the false-negative and false-positive rates using vcfeval~\citep{vcfeval}.
 
@@
 
+    % TODO 7 Mbp beta-defensin locus (chr8: 6,300,000-13,300,000)
+
+    % TODO Centromere (chr8:42,881,543-47,029,467)
+
+    % TODO Simulation with full chromosomes
+
     \subsection{Real data}
@@
 
-    We implemented a novel gap-affine pairwise aligner, \wfmash\, to accelerate the computation of the mutual
+    We implemented a novel gap-affine pairwise aligner, \wfmash\/, to accelerate the computation of the mutual
     alignment of collections of genomes, a step required for building pangenome models. We have demonstrated that

### 2021-06-14 manuscript/main.tex
added why not seed-and-extend mapping methods

@@
 %\title[short Title]{Wavefront inception enables global pairwise alignment to whole eukaryotic chromosomes}
-    \title[wfmash: whole-cromosome pairwise alignment]{Whole-chromosome pairwise alignment using a hierarchical wavefront algorithm}
+    \title[wfmash: whole-cromosome pairwise aligner]{Whole-chromosome pairwise alignment using the hierarchical wavefront algorithm}
     \author[Sample \textit{et~al}.]{
@@
         \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
-        including pangenomes building. Pangenomes are sequence models able to provide a full representation of the
-        mutual alignment of collections of genomes. The time and memory required to compute a pairwise alignment
+        including pangenome construction. Pangenomes are sequence models able to provide a full representation of the
+        mutual alignment of collections of genomes. The time and memory required to compute pairwise alignments
         increase quadratically with the sequence length, making it impractical to directly align very long sequences
-        without applying heuristic approaches to determine possible syntenic regions in the sequences. Nevertheless,
+        without applying heuristic approaches to first determine possible syntenic regions in the sequences. Nevertheless,
         with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
@@
         \\
-        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely available at
-        \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
+        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely
+        available at \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
         \\
@@
 
-    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping
-    \citep{22388286, BWA_MEM, 23103880} to variant detection~\citep{21478889}, as well as de~novo
-    assembly \citep{19251739} and pangenome building~\citep{33177663, 33066802}. Pairwise sequence alignment
-    can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary
+    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{22388286, BWA_MEM, 23103880}
+    to variant detection~\citep{21478889}, as well as de~novo assembly~\citep{19251739} and
+    pangenome construction~\citep{33177663, 33066802}. Pairwise sequence alignment can be applied to identify
+    similar regions that may indicate functional, structural, and/or evolutionary
     relationships between two biological sequences. In this context, pangenomes model the full set
@@
     relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
-    the entire genetic diversity of a population. FIXME In combination with the massive amount
-    of data available made possible thanks to the advancement in sequencing technology, presses for the
-    development of tools able to compute pairwise alignment of very long sequences.
+    the entire genetic diversity of a population. This opportunity, in combination with the massive amount of data
+    available nowadays, presses for the development of tools able to compute pairwise alignment of very long sequences.
 
-    To face the upcoming challenges in pangenomes building, we have developed \wfmash\/, a new gap-affine
-    pairwise tool for aligning DNA sequences at a whole-chromosome scale. It supports split-read alignment
-    and gap-affine penalties for insertion and deletions. \wfmash\ scales efficiently to large eukaryotic
-    chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise
-    alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the integration of
-    MashMap~\citep{30423094}, a fast approximate aligner for computing local alignment boundaries between
-    long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine algorithm
-    that takes advantage of homologous regions between the sequences to accelerate the alignment process.
+    Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory
+    when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing methods
+    like PacBio HiFi. To scale to vertebrate genomes, methods for pangenome construction based on these approaches
+    must first filter out centromeric and other highly-repetitive sequences.
 
-    %TODO Some sentences of why it exists with respect to Minimap2/Winnomap2, explaining the limitations of those tools in the pairwise alignment task.
+    In the interest of operating on the full pangenome, and to face the upcoming challenges in pangenome construction,
+    we have developed \wfmash\/, a new gap-affine pairwise tool for aligning DNA sequences at a whole-chromosome scale.
+    It supports split-read alignment and gap-affine penalties for insertion and deletions. \wfmash\ scales
+    efficiently to large eukaryotic chromosomes while requiring little time and memory, making it suitable to
+    be applied on the pairwise alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the
+    integration of MashMap~\citep{30423094}, a fast approximate sequence mapper for computing local alignment boundaries
+    between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine
+    algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
 
@@
 
-    To avoid such limitation, wfmash applies a hierarchical WFA, exploiting the WFA to guide the alignment
+    To avoid such limitation, \wfmash\ applies a hierarchical WFA, exploiting the WFA to guide the alignment
     process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
@@
 
-    We evaluated the performance of \wfmash\ 0.xx against Minimap2-v2.20-r1061 \citep{29750242} and Winnomap 2.03 \citep{XXX}, in
+    We evaluated the performance of \wfmash\ \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{29750242} and Winnomap 2.03~\citep{XXX}, in
     terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
@@
     \subsection{Simulated data}
-    We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and
-    the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of divergence and length
-    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool~\citep{splitfa},
-    In each run, all sequences have the same length, with 50\% of them in reverse complement.
-    Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced using the
-    Mutation-Simulator script~\citep{32780803}.
+    We simulated long sequences from the telomere-to-telomere (T2T) assemblies of the human chromosomes 8,
+    at different levels of length and divergence (Table 1 and Table 2). Chromosomes were split into shorter
+    sequences using the splitfa tool~\citep{splitfa}, In each run, all sequences have the same length, with 50\%
+    of them in reverse complement. Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs)
+    were introduced using the Mutation-Simulator script~\citep{32780803}.
     %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.
@@
     We implemented a novel gap-affine pairwise aligner, \wfmash\/, to accelerate the computation of the mutual
-    alignment of collections of genomes, a step required for building pangenome models. We have demonstrated that
+    alignment of collections of genomes, a step required for constructing pangenome models. We have demonstrated that
     it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
-    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, wfmash is already
+    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, \wfmash\ is already
     successfully applied in our pangenome graphs building pipeline (\citep{pggb}). No less important, pairwise

### 2021-06-14 manuscript/main.tex
a bit more explanation; citation

@@
     relationships between all the genomes represented, in contrast to reference-based approaches which
-    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
+    relate sequences to a particular genome chosen as reference. The unbiased approach allows studying
     the entire genetic diversity of a population. This opportunity, in combination with the massive amount of data
@@
     Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory
-    when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing methods
-    like PacBio HiFi. To scale to vertebrate genomes, methods for pangenome construction based on these approaches
-    must first filter out centromeric and other highly-repetitive sequences.
+    when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing
+    technologies~\citep{33288905}. To scale to vertebrate genomes, methods for pangenome construction based on these
+    approaches must first filter out centromeric and other highly-repetitive sequences, in contrast to the pangenome
+    idea of modeling the full genetic variation in the samples.
 
@@
     Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
-    runtime and memory-usage without affecting accuracy~\citep{32657365}.
+    runtime and memory-usage without affecting accuracy~\citep{32657365}. Futhermore, we made it possible
+    to set the number of mappings to return for each segment; this is useful for identifying paralogous regions and
+    homologous relationship between sets of genomes. Together with the segment length and the minimum segment
+    estimated identity, these settings allow users to precisely define the mapping space to consider,
+    specifying the characteristics of homologies to compute.
+    \\
 
-    %TODO THINGS ON THE FILTERING?
-    %TODO THING ON THE SPLIT READ AND/OR THE MERGING?
+    FIXME THINGS ON THE FILTERING???\\
 
-    In addition to the segment length, the minimum segment estimated identity and the maximum number of
-    mappings to report for each segment can be specified in the mapping process. These settings allow users
-    to precisely define the mapping space to consider, specifying the characteristics of homologies to compute.
-    Each mapping location is then used as a target for base-level alignment using a hierarchical implementation
+    FIXME THING ON THE SPLIT READ AND/OR THE MERGING???\\
+
+    Finally, each mapping location is used as a target for base-level alignment using a hierarchical implementation
     of the WFA.
@@
     \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
-    two fragments.
+    two fragments. To accelerate the process, an alignment-free comparison of the two fragments is performed first by
+    computing the mash distance between them. Cells whose fragments have a high mash distance are considered as mismatches,
+    avoiding computing the corresponding base-level alignments.
 
@@
     on the optimal path in the high order DP-matrix. During the process, a further step is performed for
-    refining the alignment. Indeed, inaccurate mapping estimation can lead to under-alignment at the beginning
-    and the end of the aligned syntenic region. Therefore, edlib is applied by computing semi-global alignments
-    to resolve the mapping boundaries.
+    refining the alignment. Inaccurate mapping estimation can lead to under-alignment at the beginning
+    and the end of the aligned syntenic region. Therefore, FIXME edlib is applied by computing semi-global alignments
+    to resolve the mapping boundaries. Moreover, the standard WFA is applied to resolve breakpoints of the structural
+    variants when present in the aligned region.
 
-    The hierarchical implementation requires only the memory to align a syntenic region at \textit{W}-bps resolution,
+    The hierarchical implementation requires only the memory to align the sequences at \textit{W}-bps resolution,
     limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
     This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
-    no heuristic seeding step, and FIXME can benefit from parallel exploration of the wavefront.
+    no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
     %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
@@
 
-    We evaluated the performance of \wfmash\ \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{29750242} and Winnomap 2.03~\citep{XXX}, in
-    terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
-    data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
-    The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K
-    CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.
+    We evaluated the performance of \wfmash\ \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{29750242} and
+    Winnomap 2.03~\citep{XXX}, in terms of runtime and memory usage. The comparison was performed on both simulated
+    and real datasets. Simulated data allowed evaluating the alignment quality by calculating precision and recall
+    in variant identification. The evaluations were performed on a workstation running Ubuntu 20.04 equipped with
+    64 GB of RAM and an Intel i9-9900K CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.
 
     \subsection{Simulated data}
-    We simulated long sequences from the telomere-to-telomere (T2T) assemblies of the human chromosomes 8,
-    at different levels of length and divergence (Table 1 and Table 2). Chromosomes were split into shorter
-    sequences using the splitfa tool~\citep{splitfa}, In each run, all sequences have the same length, with 50\%
-    of them in reverse complement. Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs)
-    were introduced using the Mutation-Simulator script~\citep{32780803}.
+    We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and the
+    telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of length and divergence
+    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool (6a0bb67)~\citep{splitfa}.
+    In each run, all sequences have the same length, with 50\% of them in reverse complement respect to the source
+    chromosome. Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced
+    using the Mutation-Simulator script~\citep{32780803}.
     %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.
 
-    To evaluate the correctness of the alignments, we used the pairwise alignment of the input sequences to build
-    a pangenome graph in each run, identifying the variants embedded in the graphs. The graphs were built using~\citep{seqwish}.
+    To evaluate the correctness of the alignments, in each run we used the pairwise alignment of the input sequences to build
+    a pangenome graph and identify the variants embedded in it. The graphs were built using~\citep{seqwish}.
     seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
     sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
-    evaluating the false-negative and false-positive rates using vcfeval~\citep{vcfeval}.
+    evaluating the false-negative and false-positive rates using vcfeval~\citep{vcfeval} for SNVs and small indels,
+    and FIXME XXXXX for SVs.
 
@@
 
-    % TODO 7 Mbp beta-defensin locus (chr8: 6,300,000-13,300,000)
+    % TODO Simulation with full chromosomes too
 
-    % TODO Centromere (chr8:42,881,543-47,029,467)
+    % TODO 7 Mbp beta-defensin locus (chr8: 6,300,000-13,300,000): wfmash/minimap2/winnomnap2
 
-    % TODO Simulation with full chromosomes
+    % TODO Centromere (chr8:42,881,543-47,029,467): only wfmash
 
     \subsection{Real data}
-    % HPRC data time/memory
+    % Yeast/HPRC data time/memory
     % Histogram of the contig length distribution (supplementary?)

### 2021-12-01 manuscript/main.tex
sync latex

@@
+\PassOptionsToPackage{utf8}{inputenc}
 \documentclass{bioinfo}
 
-\usepackage{relsize} % for Cpp
-\usepackage{xspace} % for Cpp
+\usepackage[draft]{hyperref}
+\usepackage{makecell}
+\usepackage{comment}
 
-\newcommand{\wfmash}{wfmash}
-\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
-\newcommand{\pp}{\Rplus\Rplus}
-\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}
+\usepackage{floatrow}
 
-\copyrightyear{2021} \pubyear{2021}
+% singlelinecheck=false puts subcaptions on the left
+\usepackage[singlelinecheck=false]{subcaption}
+
+\usepackage{algorithm2e}
+\usepackage[usenames,dvipsnames]{xcolor}
+
+% we squeeze our figures even more together
+\captionsetup{belowskip=-2pt}
+
+\SetAlgoLined
+\SetKwProg{MyStruct}{Struct}{ contains}{end}
+
+\newcommand{\vocab}{\textbf}
+\newcommand{\red}[1]{{\textcolor{Red}{#1}}}
+\newcommand{\FIXME}[1]{\red{[FIXME: #1]}}
+
+\def\labelitemi{--}
+
+\copyrightyear{2022} \pubyear{XXXX}
 
 \access{Advance Access Publication Date: Day Month Year}
-\appnotes{Original Paper}
+\appnotes{Sequence Analysis}
 
@@
 
-    \subtitle{Sequence analysis}
+\subtitle{Sequence Analysis}
 
-%\title[short Title]{Wavefront inception scales global pairwise alignment to whole eukaryotic chromosomes}
-%\title[short Title]{Wavefront inception enables global pairwise alignment to whole eukaryotic chromosomes}
-    \title[wfmash: whole-cromosome pairwise aligner]{Whole-chromosome pairwise alignment using the hierarchical wavefront algorithm}
-    \author[Sample \textit{et~al}.]{
-        Corresponding Author\,$^{\text{\sfb 1,}*}$,
-        Co-Author\,$^{\text{\sfb 2}}$ and
-        Co-Author\,$^{\text{\sfb 3,}*}$
-    }
-    \address{
-        $^{\text{\sf 1}}$Department, Institution, City, Post Code, Country \\
-        $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
-        $^{\text{\sf 3}}$Department, Institution, City, Post Code, Country
+\title[wfmash: whole-cromosome pairwise aligner]{wfmash: whole-chromosome pairwise alignment using the hierarchical wavefront algorithm}
+\author[Guarracino \textit{et~al}.]{
+Andrea~Guarracino\,$^{\text{\sfb 1}}$,
+Njagi~Mwaniki\,$^{\text{\sfb 2}}$,
+Santiago~ Marco-Sola\,$^{\text{\sfb 3, 4}}$,
+and~Erik~Garrison\,$^{\text{\sfb 5}*}$
 }
 
-    \corresp{$^\ast$To whom correspondence should be addressed.}
+\address{
+$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Milan, Italy \\
+$^{\text{\sf 2}}$???, ???, Pisa, Italy \\
+$^{\text{\sf 3}}$Department of Computer Sciences, Barcelona Supercomputing Center, Barcelona 08034, Spain \\
+$^{\text{\sf 4}}$Departament d’Arquitectura de Computadors i Sistemes Operatius, Universitat Autònoma de Barcelona, Barcelona 08193, Spain \\
+$^{\text{\sf 5}}$University of Tennessee Health Science Center, Memphis, TN, USA
+}
+
+\corresp{
+$^\ast$To whom correspondence should be addressed.
+%$^\dagger$Contributed equally.
+}
 
@@
 
+% XXX key message of the paper is that we have collected a set of algorithms that enable easy use of pangenome graphs for investigating biology
+
 \abstract{
-        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
-        including pangenome construction. Pangenomes are sequence models able to provide a full representation of the
-        mutual alignment of collections of genomes. The time and memory required to compute pairwise alignments
-        increase quadratically with the sequence length, making it impractical to directly align very long sequences
-        without applying heuristic approaches to first determine possible syntenic regions in the sequences. Nevertheless,
-        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
-        the development of tools able to align sequences of the order of tens of megabases long.
-        \\
-        \textbf{Results:} Here we present \wfmash\/, a new gap-affine pairwise aligner designed to align DNA sequences
-        at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
-        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
-        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
-        time and memory.
-        \\
-        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely
-        available at \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
-        \\
-        \textbf{Contact:}
-        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
-        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
-        \\
-        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
+\textbf{Motivation:}
+Pairwise alignment of sequences is an important step in many bioinformatics analyses, including pangenome construction.
+Pangenomes are sequence models able to provide a full representation of the mutual alignment of collections of genomes.
+The time and memory required to compute pairwise alignments increase quadratically with the sequence length,
+making it impractical to directly align very long sequences without applying heuristic approaches to first determine possible syntenic regions in the sequences.
+Nevertheless, with the advances in sequencing technologies, new genome assemblies are produced at a high rate,
+pressing for the development of tools able to align sequences of the order of tens of megabases long. \\
+\textbf{Results:}
+Here we present wfmash, a new gap-affine pairwise aligner designed to align DNA sequences at a whole-chromosome scale.
+wfmash applies a hierarchical implementation of the wavefront alignment algorithm to guide the alignment of very long sequences.
+It scales efficiently to large eukaryotic chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little time and memory. \\
+\textbf{Availability:}
+wfmash is published as free software under the MIT open source license.
+Source code and documentation are available at \url{https://github.com/ekg/wfmash}.
+wfmash can be installed via Bioconda \url{https://bioconda.github.io/recipes/wfmash/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/wfmash.scm}. \\
+\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
+%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
 }
@@
 \section{Introduction}
+Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{22388286, BWA_MEM, 23103880}
+to variant detection~\citep{21478889}, as well as de~novo assembly~\citep{19251739} and pangenome construction~\citep{33177663, 33066802}.
+Pairwise sequence alignment can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary relationships between two biological sequences.
+In this context, pangenomes model the full set of genomic elements in a given species or clade~\citep{32453966}.
+These data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences to a particular genome chosen as reference.
+The unbiased approach allows studying the entire genetic diversity of a population.
+This opportunity, in combination with the massive amount of data available nowadays, presses for the development of tools able to compute pairwise alignment of very long sequences.
 
-    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{22388286, BWA_MEM, 23103880}
-    to variant detection~\citep{21478889}, as well as de~novo assembly~\citep{19251739} and
-    pangenome construction~\citep{33177663, 33066802}. Pairwise sequence alignment can be applied to identify
-    similar regions that may indicate functional, structural, and/or evolutionary
-    relationships between two biological sequences. In this context, pangenomes model the full set
-    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
-    relationships between all the genomes represented, in contrast to reference-based approaches which
-    relate sequences to a particular genome chosen as reference. The unbiased approach allows studying
-    the entire genetic diversity of a population. This opportunity, in combination with the massive amount of data
-    available nowadays, presses for the development of tools able to compute pairwise alignment of very long sequences.
-
-    Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory
-    when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing
-    technologies~\citep{33288905}. To scale to vertebrate genomes, methods for pangenome construction based on these
-    approaches must first filter out centromeric and other highly-repetitive sequences, in contrast to the pangenome
-    idea of modeling the full genetic variation in the samples.
+Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory when generating
+sensitive alignments for chromosome-scale contigs available thanks to the new sequencing technologies~\citep{33288905}.
+To scale to vertebrate genomes, methods for pangenome construction based on these approaches must first filter out centromeric
+and other highly-repetitive sequences, in contrast to the pangenome idea of modeling the full genetic variation in the samples.
 
 In the interest of operating on the full pangenome, and to face the upcoming challenges in pangenome construction,
-    we have developed \wfmash\/, a new gap-affine pairwise tool for aligning DNA sequences at a whole-chromosome scale.
-    It supports split-read alignment and gap-affine penalties for insertion and deletions. \wfmash\ scales
-    efficiently to large eukaryotic chromosomes while requiring little time and memory, making it suitable to
-    be applied on the pairwise alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the
-    integration of MashMap~\citep{30423094}, a fast approximate sequence mapper for computing local alignment boundaries
-    between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine
-    algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
-
-%\enlargethispage{12pt}
+we have developed wfmash, a new gap-affine pairwise tool for aligning DNA sequences at a whole-chromosome scale.
+It supports split-read alignment and gap-affine penalties for insertion and deletions.
+wfmash scales efficiently to large eukaryotic chromosomes while requiring little time and memory, making it suitable to
+be applied on the pairwise alignment of hundreds of genomes.
+wfmash is a DNA sequence aligner based on the integration of MashMap~\citep{30423094}, a fast approximate sequence mapper
+for computing local alignment boundaries between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952},
+an exact gap-affine algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
 
@@
 
-    \wfmash\ first applies a locality-sensitive hashing, from MashMap, to rapidly determine syntenic region
+wfmash first applies a locality-sensitive hashing, from MashMap, to rapidly determine syntenic region
 boundaries between long DNA sequences. Then, a hierarchical implementation of the WFA allows computing the
@@
 \subsection{Approximate mapping}
-
-    Each query sequence is broken into non-overlapping pieces of the requested length. These segments are then
-    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, incorporating
-    the robust winnowing~\citep{Schleimer S. et al. (2003)} in the minimizer downsampling.
-    Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
-    runtime and memory-usage without affecting accuracy~\citep{32657365}. Futhermore, we made it possible
-    to set the number of mappings to return for each segment; this is useful for identifying paralogous regions and
-    homologous relationship between sets of genomes. Together with the segment length and the minimum segment
-    estimated identity, these settings allow users to precisely define the mapping space to consider,
-    specifying the characteristics of homologies to compute.
+Each query sequence is broken into non-overlapping pieces of the requested length.
+These segments are then mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}.
+We extended MashMap, incorporating the robust winnowing~\citep{Schleimer S. et al. (2003)} in the minimizer downsampling.
+Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in runtime and memory-usage without affecting accuracy~\citep{32657365}.
+Futhermore, we made it possible to set the number of mappings to return for each segment; this is useful for identifying
+paralogous regions and homologous relationship between sets of genomes.
+Together with the segment length and the minimum segment estimated identity, these settings allow users to precisely
+define the mapping space to consider, specifying the characteristics of homologies to compute.
 \\
-
-    FIXME THINGS ON THE FILTERING???\\
-
-    FIXME THING ON THE SPLIT READ AND/OR THE MERGING???\\
+FIXME THINGS ON THE FILTERING???
+\\
+FIXME THING ON THE SPLIT READ AND/OR THE MERGING???
+\\
 
@@
 \subsection{Base-level alignment}
+The time and memory required to compute the base-level alignment increases quadratically with the sequence length.
+The WFA provides an efficient way to decrease the amount of computation required to obtain the optimal alignment between two sequences,
+reducing the cost to be quadratic in the alignment penalty score of the optimal alignment~\citep{32915952}.
+This means that the algorithm is very efficient in aligning similar homologous sequences (i.e., sequences with a
+low alignment penalty score), but high divergence genomes and/or noisy long reads can increase its memory and runtime costs.
 
-    The time and memory required to compute the base-level alignment increases quadratically with the sequence
-    length. The WFA provides an efficient way to decrease the amount of computation required to obtain the
-    optimal alignment between two sequences, reducing the cost to be quadratic in the alignment penalty score
-    of the optimal alignment~\citep{32915952}. This means that the algorithm is very efficient in aligning
-    similar homologous sequences (i.e., sequences with a low alignment penalty score), but high divergence
-    genomes and/or noisy long reads can increase its memory and runtime costs.
+To avoid such limitation, wfmash applies a hierarchical WFA, exploiting the WFA to guide the alignment process, but keeping the largest alignment problem size small.
+Rather than directly aligning the whole sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment (Figure~1\vphantom{\ref{fig:1}}).
+The whole global alignment is then computed over the full dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
+Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments \textit{W}-bps long from the two sequences to align.
+To determine if each cell is a match, and then guiding the \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the two fragments.
+To accelerate the process, an alignment-free comparison of the two fragments is performed first by computing the mash distance between them.
+Cells whose fragments have a high mash distance are considered as mismatches, avoiding computing the corresponding base-level alignments.
 
-    To avoid such limitation, \wfmash\ applies a hierarchical WFA, exploiting the WFA to guide the alignment
-    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
-    sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment
-    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
-    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
-    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
-    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
-    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
-    two fragments. To accelerate the process, an alignment-free comparison of the two fragments is performed first by
-    computing the mash distance between them. Cells whose fragments have a high mash distance are considered as mismatches,
-    avoiding computing the corresponding base-level alignments.
-
-    \begin{figure}[!tpb]%figure1
-        \fboxsep=0pt\colorbox{gray}{
-            \begin{minipage}[t]{235pt}
-                \vbox to 100pt{\vfill\hbox to
-                235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}\vfill}
-            \end{minipage}}
-        %\centerline{\includegraphics{fig01.eps}}
-        \caption{Generate a cool plot using R as a base to create a cool image.}\label{fig:1}
-    \end{figure}
-
-%\begin{figure}[!tpb]%figure2
-%%\centerline{\includegraphics{fig02.eps}}
-%\caption{Caption, caption.}\label{fig:02}
-%\end{figure}
-
-    Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments
-    on the optimal path in the high order DP-matrix. During the process, a further step is performed for
-    refining the alignment. Inaccurate mapping estimation can lead to under-alignment at the beginning
-    and the end of the aligned syntenic region. Therefore, FIXME edlib is applied by computing semi-global alignments
-    to resolve the mapping boundaries. Moreover, the standard WFA is applied to resolve breakpoints of the structural
-    variants when present in the aligned region.
+Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments on the optimal path in the high order DP-matrix.
+During the process, a further step is performed for refining the alignment.
+Inaccurate mapping estimation can lead to under-alignment at the beginning and the end of the aligned syntenic region.
+Therefore, WFA is applied by computing semi-global alignments to resolve the mapping boundaries.
+Moreover, the standard WFA is applied to resolve breakpoints of the structural variants when present in the aligned region.
 
@@
 
+\input{fig_sorting}
+
 \section{Results}
 
-    We evaluated the performance of \wfmash\ \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{29750242} and
+We evaluated the performance of wfmash \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{29750242} and
 Winnomap 2.03~\citep{XXX}, in terms of runtime and memory usage. The comparison was performed on both simulated
@@
 telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of length and divergence
-    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool (6a0bb67)~\citep{splitfa}.
+(\FIXME{MAYBETable 1 and Table 2}). Chromosomes were split into shorter sequences using the splitfa tool (6a0bb67)~\citep{splitfa}.
 In each run, all sequences have the same length, with 50\% of them in reverse complement respect to the source
@@
 
+\begin{comment}
 \begin{table}[!t]
@@
 \end{table}
+\end{comment}
 
@@
 \subsection{Real data}
+XXXXX
+\\
 % Yeast/HPRC data time/memory
@@
 \section{Discussion}
-
-    We implemented a novel gap-affine pairwise aligner, \wfmash\/, to accelerate the computation of the mutual
+We implemented a novel gap-affine pairwise aligner, wfmash, to accelerate the computation of the mutual
 alignment of collections of genomes, a step required for constructing pangenome models. We have demonstrated that
 it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
-    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, \wfmash\ is already
-    successfully applied in our pangenome graphs building pipeline (\citep{pggb}). No less important, pairwise
+Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, wfmash is already
+successfully applied in our pangenome graphs building pipeline~\citep{pggb}. No less important, pairwise
 alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the
@@
 
-    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
 
-    \vspace*{+12pt}
-    \textit{Financial Support}: none declared.
+We thank members of the HPRC Pangenome Working Group for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
 
-    \textit{Conflict of Interest}: none declared.
-    \vspace*{+12pt}
+\section*{Funding}
+
+We gratefully acknowledge support from NIH/NIDA U01DA047638 and NIH/NIGMS R01GM123489 (EG).
 
@@
 
-    Data used for the simulations is available at \href{https://github.com/marbl/CHM13}{https://github.com/marbl/CHM13}.
-    Data used for Human pangenome graph analysis is available at \href{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}{https://github.com/human-pangenomics/HPP\_Year1\_Data\_Freeze\_v1.0}.
+%Data used to build human pangenome graphs: \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}. 
+%The graphs we generated the figures from: \url{https://s3-us-west-2.amazonaws.com/human-pangenomics/index.html?prefix=pangenomes/scratch/2021_11_04_pggb_wgg.87/}.
+Code and links to data resources used to build this manuscript and its figures, can be found in the paper's public repository: \url{https://github.com/AndreaGuarracino/wfmash-paper}.
 
-    \section*{Acknowledgements}
-    %ToDo
-    XXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
-
-    \section*{Funding}
-    %ToDo
-    This work has been supported by the XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
-
-%\bibliographystyle{natbib}
+\bibliographystyle{natbib}
 %\bibliographystyle{achemnat}
@@
 %
-%\bibliography{Document}
+\bibliography{document}
 
-    \begin{thebibliography}{}
+% \begin{thebibliography}{}
 
-        \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-        Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
+% \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
+% Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
 
-        \bibitem[Bag {\it et~al}., 2001]{Bag01}
-        Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
+% \bibitem[Bag {\it et~al}., 2001]{Bag01}
+% Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
 
-        \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-        Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-        in nigral dopaminergic neurnol cell: correlation with the known
-        pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-        Res.}, \textbf{110}(Suppl. 1), 76--84.
+% \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
+% Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
+% in nigral dopaminergic neurnol cell: correlation with the known
+% pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
+% Res.}, \textbf{110}(Suppl. 1), 76--84.
 
-        \bibitem[Lehmann, 1986]{Leh86}
-        Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
+% \bibitem[Lehmann, 1986]{Leh86}
+% Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
 
-        \bibitem[Crenshaw and Jones, 2003]{Cre03}
-        Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-        cancer management: one tumor, one chip. \textit{Bioinformatics},
-        doi:10.1093/bioinformatics/btn000.
+% \bibitem[Crenshaw and Jones, 2003]{Cre03}
+% Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
+% cancer management: one tumor, one chip. \textit{Bioinformatics},
+% doi:10.1093/bioinformatics/btn000.
 
-        \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-        Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-        (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-        ???--???.
+% \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
+% Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
+% (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
+% ???--???.
 
-        \bibitem[Bardet, 1920]{Bar20}
-        Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-        polydactylie et retinite pigmentaire (contribution a l'etude des
-        formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-        institution, Paris, France.
+% \bibitem[Bardet, 1920]{Bar20}
+% Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
+% polydactylie et retinite pigmentaire (contribution a l'etude des
+% formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
+% institution, Paris, France.
 
-    \end{thebibliography}
+% \end{thebibliography}
 \end{document}

### 2021-12-02 scripts/simulations.sh
unlock minimap2 and winnomap2

@@ -202,7 +202,7 @@ do
 
 	for aligner in wfmash; do mkdir -p graphs/$aligner; for idx_len in ${!lengths[*]}; do len=${lengths[$idx_len]}; path_input_fa_gz=seqs/${name_input_fasta}+samples_$divergence.l${len}.fa.gz; path_input_paf=alignments/$aligner/${name_input_fasta}+samples_$divergence.l${len}.paf; path_gfa=graphs/$aligner/${name_input_fasta}+samples_$divergence.l${len}.gfa; echo $path_gfa; seqwish -t $threads -s $path_input_fa_gz -p $path_input_paf -k 0 -g $path_gfa -B 9000000000; done; done;
 
-	#for aligner in minimap2 winnomap2; do mkdir -p graphs/$aligner; for idx_len in ${!lengths[*]}; do len=${lengths[$idx_len]}; path_input_fa_gz=seqs/${name_input_fasta}+samples_$divergence.l${len}.fa.gz; for l in 0 100000 400000 800000; do path_input_paf=alignments/$aligner/${name_input_fasta}+samples_$divergence.l${len}.drop$l.paf; path_gfa=graphs/$aligner/${name_input_fasta}+samples_$divergence.l${len}.drop$l.gfa; echo $path_gfa; seqwish -t $threads -s $path_input_fa_gz -p $path_input_paf -k 0 -g $path_gfa -B 1000000000; done; done; done;
+	for aligner in minimap2 winnomap2; do mkdir -p graphs/$aligner; for idx_len in ${!lengths[*]}; do len=${lengths[$idx_len]}; path_input_fa_gz=seqs/${name_input_fasta}+samples_$divergence.l${len}.fa.gz; for l in 0 100000 400000 800000; do path_input_paf=alignments/$aligner/${name_input_fasta}+samples_$divergence.l${len}.drop$l.paf; path_gfa=graphs/$aligner/${name_input_fasta}+samples_$divergence.l${len}.drop$l.gfa; echo $path_gfa; seqwish -t $threads -s $path_input_fa_gz -p $path_input_paf -k 0 -g $path_gfa -B 1000000000; done; done; done;
 done
 
 
@@ -219,7 +219,7 @@ do
 
[…]

### 2021-12-04 manuscript/main.tex
add affiliation; start simulations workflow for Octopus

@@
 $^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Milan, Italy \\
-$^{\text{\sf 2}}$???, ???, Pisa, Italy \\
+$^{\text{\sf 2}}$Department of Computer Sciences, University of Pisa, Pisa, Italy \\
 $^{\text{\sf 3}}$Department of Computer Sciences, Barcelona Supercomputing Center, Barcelona 08034, Spain \\

### 2021-12-04 scripts/simulations.sh

@@ -27,45 +27,45 @@ run_meryl=~/git/Winnowmap/bin/meryl
 # Input
 assembly='CHM13_v1.1'
 species='Homo sapiens'
-path_input_fasta=chr20.fa
-path_input_sdf=chr20.sdf
+path_input_fasta=chm13.draft_v1.1.fa
+path_input_sdf=chm13.draft_v1.1.sdf
 
 
 # Variables
 header_input_fasta=$(grep '>' $path_input_fasta | sed 's/>//g')
 name_input_fasta=$(basename $path_input_fasta .fa)
 len_input_fasta=$(cut $path_input_fasta.fai -f2)
-threads=16
+threads=14
 
 divergences=(
 #  0.001
-  0.01
-#  0.05
+#  0.01
+  0.05
 #  0.15
 #  0.20
 )
 
 presets=(
-  'asm5'
-#  'asm10'
+#  'asm5'
+  'asm10'
 #  'asm20'
 )
 
 lengths=(
 	#200000
 	#500000
-	1000000
-	#5000000
+	#1000000
+	5000000
 	#30000000
 	#70000000
 	#150000000
 )
 
-samples=({A..C})
+samples=({A..D})
 
 
 # Mutated chromosome generation (SNVs + small INDELs)
-variant_block=5
+variant_block=10
 
 mkdir -p chromosomes
 for index in ${!divergences[*]};
@@ -80,7 +80,7 @@ do
 
 	name_output_pre=${name_input_fasta}_${divergence}_pre
 	prefix_mutations_vcf_pre=chromosomes/$name_output_pre
-	python3 $path_mutation_simulator_py $path_input_fasta --output $prefix_mutations_vcf_pre args --snp $snv_rate -titv 2 --insert $ins_rate --snpblock $variant_block --insertlength 2000 --insertblock $variant_block --deletion $del_rate --deletionlength 2000 --deletionblock $variant_block --assembly $assembly --species "$species" --sample ${name_input_fasta}_mt
+	python3 $path_mutation_simulator_py $path_input_fasta --output $prefix_mutations_vcf_pre args --snp $snv_rate -titv 2 --insert $ins_rate --snpblock $variant_block --insertlength 5 --insertblock $variant_block --deletion $del_rate --deletionlength 5 --deletionblock $variant_block --assembly $assembly --species "$species" --sample ${name_input_fasta}_mt
 
[…]

### 2022-02-17 scripts/minihprc_test/fasta+paf2vcf.mod.sh
added mini HPRC test

@@ -1,0 +1,35 @@
+#!/bin/bash
+
+# Example:
+# bash scripts/fasta+paf2vcf.sh data/LPA.subset.fa.gz 100k 300k 98 14 chm13 '_' ~/tools/wfmash/build/bin/wfmash-master master
+
+# Inputs
+FASTA=$1
+SEGMENT_LEN=$2
+BLOCK_LEN=$3
+IDENTITY=$4
+NUM_HAPLOTYPES=$5
+REFERENCE_PREFIX=$6
+SEPARATOR=$7
+PATH_WFMASH=$8
+OUTPUT_PREFIX=$9
+
+FASTA_PREFIX=$(basename $FASTA .fa.gz)
+
+PATH_SEQWISH=~/tools/seqwish/bin/seqwish-ccfefb016fcfc9937817ce61dc06bbcf382be75e
+PATH_VG=~/tools/vg
+
+# Paths
+PAF=${OUTPUT_PREFIX}.$FASTA_PREFIX.s${SEGMENT_LEN}.l${BLOCK_LEN}.p$IDENTITY.n${NUM_HAPLOTYPES}.k16.paf
+GFA=$PAF.k0.B10M.gfa
+VCF=$GFA.vcf.gz
+
+echo "All-vs-all alignment"
+\time -v $PATH_WFMASH $FASTA $FASTA -X -s ${SEGMENT_LEN} -l ${BLOCK_LEN} -p $IDENTITY -n ${NUM_HAPLOTYPES} -k 16 -t 48 > $PAF
+
+echo "Graph induction"
+# -k 0` is for getting a lossless representation of the pairwise alignment
+\time -v ${PATH_SEQWISH} -s $FASTA -p $PAF -g $GFA -k 0 -B 10M -t 48 -P
+
+echo "Identify variants"
+\time -v ${PATH_VG} deconstruct -e -a -P $REFERENCE_PREFIX -H $SEPARATOR $GFA -t 48 | bgzip -c > $VCF && tabix $VCF

### 2022-02-17 scripts/minihprc_test/mini_hprc.sh

@@ -1,0 +1,45 @@
+#!/bin/bash
+
+PREFIX=$1
+PATH_WFMASH=$2
+
+DIR_FASTA=/lizardfs/guarracino/HPRC/mini_dataset
+DIR_REGIONS=/lizardfs/guarracino/HPRC/mini_dataset/union
+OUTPUT_DIR=/lizardfs/guarracino/HPRC/mini_dataset
+PATH_FASTA_PAF_TO_VCF=/lizardfs/guarracino/HPRC/mini_dataset/fasta+paf2vcf.mod.sh
+PATH_VCF_PREPROCESS_SH=/lizardfs/guarracino/HPRC/mini_dataset/vcf_preprocess.sh
+PATH_VCF_EVALUATION_SH=/lizardfs/guarracino/HPRC/mini_dataset/vcf_evaluation.sh
+
+
+( seq 22; echo X; echo Y; echo M ) | while read i; do
+        #echo $i
+
+        if [[ ! -s ${DIR_REGIONS}/GRCh38_notinalldifficultregions.chr$i.bed.gz ]]; then
+                zgrep -P "^chr$i\t" ${DIR_REGIONS}/GRCh38_notinalldifficultregions.bed.gz | bgzip > ${DIR_REGIONS}/GRCh38_notinalldifficultregions.chr$i.bed.gz;
+        fi
+        if [[ ! -s ${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz ]]; then
+                zgrep -P "^chr$i\t" ${DIR_REGIONS}/GRCh38_alldifficultregions.bed.gz | bgzip > ${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz;
+        fi
+
+        OUTPUT_DIR=${OUTPUT_DIR}/HG00438_$PREFIX/chr${i}
+
+        if [ ! -d ${OUTPUT_DIR} ]; then
+                #echo $PREFIX
+
+                mkdir -p $OUTPUT_DIR
+
+                FASTA=${DIR_FASTA}/chr${i}hg00.fa.gz
+                PATH_EASY_REGIONS=${DIR_REGIONS}/GRCh38_notinalldifficultregions.chr$i.bed.gz
+                PATH_HARD_REGIONS=${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz
+
[…]

### 2022-02-17 scripts/minihprc_test/single_hprc_chr.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+
+FASTA=$1
+PREFIX=$2
+PATH_WFMASH=$3
+OUTPUT_DIR=$4
+PATH_EASY_REGIONS=$5
+PATH_HARD_REGIONS=$6
+PATH_FASTA_PAF_TO_VCF=$7
+PATH_VCF_PREPROCESS_SH=$8
+PATH_VCF_EVALUATION_SH=$9
+
+cd /scratch
+bash ${PATH_FASTA_PAF_TO_VCF} $FASTA 100k 300k 98 14 grch38 '#' ${PATH_WFMASH} ${PREFIX};
+mv ${PREFIX}* ${OUTPUT_DIR}
+
+FASTA_PREFIX=$(basename $FASTA .fa.gz)
+VCF=${OUTPUT_DIR}/$PREFIX.$FASTA_PREFIX.s100k.l300k.p98.n14.k16.paf.k0.B10M.gfa.vcf.gz
+
+cd ${OUTPUT_DIR}
+bash ${PATH_VCF_EVALUATION_SH} HG00438 $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} ${OUTPUT_DIR} ${PATH_VCF_PREPROCESS_SH}

### 2022-02-17 scripts/minihprc_test/statistics.sh

@@ -1,0 +1,54 @@
+prefix1=2f9e6a889
+prefix2=wflign-div-bffbe87
+
+
+# Runtimes
+
+# Memory
+grep 'Command\|Maximum' HG00438_*/sl* | grep $prefix1 | grep wfmash -A 1 | grep Maximum | cut -f 1,3 -d ':' > tmp1
+grep 'Command\|Maximum' HG00438_*/sl* | grep $prefix2 | grep wfmash -A 1 | grep Maximum | cut -f 1,3 -d ':' > tmp2
+#paste tmp1 tmp2 | awk '{print($1,($4-$2)/1024/1024)}' | column -t # GB
+paste tmp1 tmp2 | awk '{print($1,($2/$4))}' | column -t
+rm tmp1 tmp2
+
+# Num. matching bases
+ls HG00438_*/*/*paf | grep $prefix1 | while read f; do nmatches=$(< $f awk '{ sum += $10 } END { print sum }'); echo "$f: $nmatches" >> tmp1; done
+ls HG00438_*/*/*paf | grep $prefix2 | while read f; do nmatches=$(< $f awk '{ sum += $10 } END { print sum }'); echo "$f: $nmatches" >> tmp2; done
+paste tmp1 tmp2 | awk '{print($1,$2-$4)}' | cut -f 3 -d '/' | column -t
+rm tmp1 tmp2
+
+# F1-scores
+paste \
+	<(grep None */*/*/*txt | grep $prefix1 | tr -s ' ' | cut -f 1,9 -d ' ') \
+	<(grep None */*/*/*txt | grep $prefix2 | tr -s ' ' | cut -f 1,9 -d ' ') |\
+	awk '{print($1, $2-$4)}'
+
+
+
+
+# Num. matching bases (wfa2-integration - master-2f9e6a8)
+2f9e6a889809b6cb90aae5c06e896db0c59e23c5.chr1hg00.s100k.l300k.p98.n14.k16.paf:   1681514
+2f9e6a889809b6cb90aae5c06e896db0c59e23c5.chr10hg00.s100k.l300k.p98.n14.k16.paf:  949830
+2f9e6a889809b6cb90aae5c06e896db0c59e23c5.chr11hg00.s100k.l300k.p98.n14.k16.paf:  287760
[…]

### 2022-02-17 scripts/minihprc_test/vcf_evaluation.sh

@@ -1,0 +1,43 @@
+#!/bin/bash
+
+# Variables
+REF=/lizardfs/erikg/HPRC/year1v2genbank/evaluation/GCA_000001405.15_GRCh38_no_alt_analysis_set.fna
+REF_SDF=/lizardfs/erikg/HPRC/year1v2genbank/evaluation/GCA_000001405.15_GRCh38_no_alt_analysis_set.fna.sdf
+TRUTH_VCF=/lizardfs/erikg/HPRC/year1v2genbank/evaluation/HG00438.GRCh38_no_alt.deepvariant.vcf.gz
+
+# Inputs
+SAMPLE=$1
+QUERY_VCF=$2
+EASY_REGIONS_BED=$3
+HARD_REGIONS_BED=$4
+OUTPUT_DIR=$5
+PATH_VCF_PREPROCESS_SH=$6
+
+echo "VCF renaming"
+zcat $QUERY_VCF | sed 's/^grch38#//g' | bgzip -c -@ 48 > "$QUERY_VCF".renamed.vcf.gz && tabix "$QUERY_VCF".renamed.vcf.gz
+
+echo "VCF processing"
+bash ${PATH_VCF_PREPROCESS_SH} "$QUERY_VCF".renamed.vcf.gz "$SAMPLE" 50
+rm "$QUERY_VCF".renamed.vcf.gz
+
+FNAME=$(dirname "$QUERY_VCF")/$(basename "$QUERY_VCF".renamed.vcf.gz)
+PREFIX="${FNAME%.vcf.gz}"
+NORMALIZED_VCF=${PREFIX}.max50.chr1-22.vcf.gz
+
+echo "VCF evaluation"
+
+/gnu/store/3vmp4dw8y0r49h0hbjbgv3bckgvz4k0m-rtg-tools-3.11/rtg vcfeval \
+    -t "$REF_SDF" \
+    -b "$TRUTH_VCF" \
+    -e "$EASY_REGIONS_BED" \
+    -c "$NORMALIZED_VCF" \
+    -T 48 \
+    -o "$OUTPUT_DIR"/easy
+
+/gnu/store/3vmp4dw8y0r49h0hbjbgv3bckgvz4k0m-rtg-tools-3.11/rtg vcfeval \
+    -t "$REF_SDF" \
+    -b "$TRUTH_VCF" \
+    -e "$HARD_REGIONS_BED" \
+    -c "$NORMALIZED_VCF" \
+    -T 48 \
+    -o "$OUTPUT_DIR"/hard

### 2022-02-17 scripts/minihprc_test/vcf_preprocess.sh

@@ -1,0 +1,26 @@
+#!/bin/bash
+
+# From https://github.com/wwliao/pangenome-utils/blob/main/preprocess_vcf.sh
+# Usage: preprocess_vcf.sh <VCF file> <sample name> <max variant size>
+
+VCF=$1
+FNAME=$(basename $VCF)
+PREFIX=$(dirname $VCF)/"${FNAME%.vcf.gz}"
+SAMPLE=$2
+CHROMS="chr1,chr2,chr3,chr4,chr5,chr6,chr7,chr8,chr9,chr10,chr11,chr12,chr13,chr14,chr15,chr16,chr17,chr18,chr19,chr20,chr21,chr22,chrX,chrY"
+MAXSIZE=$3
+REF="/lizardfs/erikg/HPRC/year1v2genbank/evaluation/GCA_000001405.15_GRCh38_no_alt_analysis_set.fna"
+MEM="10G"
+
+bcftools=/gnu/store/pr0v7dvbf16qkm45fiw26hspy77w5bb2-bcftools-1.12/bin/bcftools
+$bcftools view -a -s ${SAMPLE} -Ou ${VCF} \
+    | $bcftools norm -f ${REF} -c s -m - -Ou \
+    | $bcftools view -e 'GT="ref" | GT~"\."' -f 'PASS,.' -Ou \
+    | $bcftools sort -m ${MEM} -T bcftools-sort.XXXXXX -Ou \
+    | $bcftools norm -d exact -Oz -o ${PREFIX}.norm.vcf.gz \
+    && $bcftools index -t ${PREFIX}.norm.vcf.gz \
+    && $bcftools view -e "STRLEN(REF)>${MAXSIZE} | STRLEN(ALT)>${MAXSIZE}" \
+                 -r ${CHROMS} -Oz -o ${PREFIX}.max${MAXSIZE}.chr1-22.vcf.gz \
+                 ${PREFIX}.norm.vcf.gz \
+    && $bcftools index -t ${PREFIX}.max${MAXSIZE}.chr1-22.vcf.gz \
+    && rm ${PREFIX}.norm.vcf.gz*

### 2022-02-17 scripts/minihprc_test/mini_hprc.sh
multi sample mini HPRC test

@@ -5,6 +5,10 @@ PATH_WFMASH=$2
 
 DIR_FASTA=/lizardfs/guarracino/HPRC/mini_dataset
 DIR_REGIONS=/lizardfs/guarracino/HPRC/mini_dataset/union
+
+# From https://s3-us-west-2.amazonaws.com/human-pangenomics/index.html?prefix=submissions/B581EBA7-8BDE-4C7C-9DEA-78B99A051155--Yale_HPP_Year1_Variant_Calls/
+DIR_TRUTH_VCF=/lizardfs/guarracino/HPRC/mini_dataset/truth/
+
 OUTPUT_DIR=/lizardfs/guarracino/HPRC/mini_dataset
 PATH_FASTA_PAF_TO_VCF=/lizardfs/guarracino/HPRC/mini_dataset/fasta+paf2vcf.mod.sh
 PATH_VCF_PREPROCESS_SH=/lizardfs/guarracino/HPRC/mini_dataset/vcf_preprocess.sh
@@ -21,7 +25,7 @@ PATH_VCF_EVALUATION_SH=/lizardfs/guarracino/HPRC/mini_dataset/vcf_evaluation.sh
                 zgrep -P "^chr$i\t" ${DIR_REGIONS}/GRCh38_alldifficultregions.bed.gz | bgzip > ${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz;
         fi
 
-        OUTPUT_DIR=${OUTPUT_DIR}/HG00438_$PREFIX/chr${i}
+        OUTPUT_DIR=${OUTPUT_DIR}/$PREFIX/chr${i}
 
         if [ ! -d ${OUTPUT_DIR} ]; then
                 #echo $PREFIX
@@ -32,7 +36,7 @@ PATH_VCF_EVALUATION_SH=/lizardfs/guarracino/HPRC/mini_dataset/vcf_evaluation.sh
                 PATH_EASY_REGIONS=${DIR_REGIONS}/GRCh38_notinalldifficultregions.chr$i.bed.gz
                 PATH_HARD_REGIONS=${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz
 
[…]

### 2022-02-17 scripts/minihprc_test/single_hprc_chr.sh

@@ -7,8 +7,9 @@ OUTPUT_DIR=$4
 PATH_EASY_REGIONS=$5
 PATH_HARD_REGIONS=$6
 PATH_FASTA_PAF_TO_VCF=$7
-PATH_VCF_PREPROCESS_SH=$8
-PATH_VCF_EVALUATION_SH=$9
+DIR_TRUTH_VCF=$8
+PATH_VCF_PREPROCESS_SH=$9
+PATH_VCF_EVALUATION_SH=${10}
 
 cd /scratch
 bash ${PATH_FASTA_PAF_TO_VCF} $FASTA 100k 300k 98 14 grch38 '#' ${PATH_WFMASH} ${PREFIX};
@@ -17,5 +18,10 @@ mv ${PREFIX}* ${OUTPUT_DIR}
 FASTA_PREFIX=$(basename $FASTA .fa.gz)
 VCF=${OUTPUT_DIR}/$PREFIX.$FASTA_PREFIX.s100k.l300k.p98.n14.k16.paf.k0.B10M.gfa.vcf.gz
 
-cd ${OUTPUT_DIR}
-bash ${PATH_VCF_EVALUATION_SH} HG00438 $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} ${OUTPUT_DIR} ${PATH_VCF_PREPROCESS_SH}
+for SAMPLE in HG00438 HG00621 HG00673 HG00733 HG00735 HG00741; do
+  TRUTH_VCF=${DIR_TRUTH_VCF}/$SAMPLE.GRCh38_no_alt.deepvariant.vcf.gz
+
+  cd ${OUTPUT_DIR}
+  bash ${PATH_VCF_EVALUATION_SH} $SAMPLE $TRUTH_VCF $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} ${OUTPUT_DIR}/$SAMPLE ${PATH_VCF_PREPROCESS_SH}
+done
+

### 2022-02-17 scripts/minihprc_test/vcf_evaluation.sh

@@ -3,15 +3,15 @@
 # Variables
 REF=/lizardfs/erikg/HPRC/year1v2genbank/evaluation/GCA_000001405.15_GRCh38_no_alt_analysis_set.fna
 REF_SDF=/lizardfs/erikg/HPRC/year1v2genbank/evaluation/GCA_000001405.15_GRCh38_no_alt_analysis_set.fna.sdf
-TRUTH_VCF=/lizardfs/erikg/HPRC/year1v2genbank/evaluation/HG00438.GRCh38_no_alt.deepvariant.vcf.gz
 
 # Inputs
 SAMPLE=$1
-QUERY_VCF=$2
-EASY_REGIONS_BED=$3
-HARD_REGIONS_BED=$4
-OUTPUT_DIR=$5
-PATH_VCF_PREPROCESS_SH=$6
+TRUTH_VCF=$2
+QUERY_VCF=$3
+EASY_REGIONS_BED=$4
+HARD_REGIONS_BED=$5
+OUTPUT_DIR=$6
+PATH_VCF_PREPROCESS_SH=$7
 
 echo "VCF renaming"
 zcat $QUERY_VCF | sed 's/^grch38#//g' | bgzip -c -@ 48 > "$QUERY_VCF".renamed.vcf.gz && tabix "$QUERY_VCF".renamed.vcf.gz

### 2022-02-17 scripts/minihprc_test/fasta+paf2vcf.mod.sh
speed up the mini HPRC test

@@ -12,7 +12,7 @@ NUM_HAPLOTYPES=$5
 REFERENCE_PREFIX=$6
 SEPARATOR=$7
 PATH_WFMASH=$8
-OUTPUT_PREFIX=$9
+PREFIX=$9
 
 FASTA_PREFIX=$(basename $FASTA .fa.gz)
 
@@ -20,7 +20,7 @@ PATH_SEQWISH=~/tools/seqwish/bin/seqwish-ccfefb016fcfc9937817ce61dc06bbcf382be75
 PATH_VG=~/tools/vg
 
 # Paths
-PAF=${OUTPUT_PREFIX}.$FASTA_PREFIX.s${SEGMENT_LEN}.l${BLOCK_LEN}.p$IDENTITY.n${NUM_HAPLOTYPES}.k16.paf
+PAF=${PREFIX}.$FASTA_PREFIX.s${SEGMENT_LEN}.l${BLOCK_LEN}.p$IDENTITY.n${NUM_HAPLOTYPES}.k16.paf
 GFA=$PAF.k0.B10M.gfa
 VCF=$GFA.vcf.gz
 

### 2022-02-17 scripts/minihprc_test/single_hprc_chr.sh

@@ -13,15 +13,14 @@ PATH_VCF_EVALUATION_SH=${10}
 
 cd /scratch
 bash ${PATH_FASTA_PAF_TO_VCF} $FASTA 100k 300k 98 14 grch38 '#' ${PATH_WFMASH} ${PREFIX};
-mv ${PREFIX}* ${OUTPUT_DIR}
 
 FASTA_PREFIX=$(basename $FASTA .fa.gz)
-VCF=${OUTPUT_DIR}/$PREFIX.$FASTA_PREFIX.s100k.l300k.p98.n14.k16.paf.k0.B10M.gfa.vcf.gz
+VCF=$PREFIX.$FASTA_PREFIX.s100k.l300k.p98.n14.k16.paf.k0.B10M.gfa.vcf.gz
 
 for SAMPLE in HG00438 HG00621 HG00673 HG00733 HG00735 HG00741; do
   TRUTH_VCF=${DIR_TRUTH_VCF}/$SAMPLE.GRCh38_no_alt.deepvariant.vcf.gz
 
-  cd ${OUTPUT_DIR}
   bash ${PATH_VCF_EVALUATION_SH} $SAMPLE $TRUTH_VCF $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} ${OUTPUT_DIR}/$SAMPLE ${PATH_VCF_PREPROCESS_SH}
 done
 
+mv ${PREFIX} ${OUTPUT_DIR}

### 2022-02-17 scripts/minihprc_test/single_hprc_chr.sh
fix output dir

@@ -20,7 +20,7 @@ VCF=$PREFIX.$FASTA_PREFIX.s100k.l300k.p98.n14.k16.paf.k0.B10M.gfa.vcf.gz
 for SAMPLE in HG00438 HG00621 HG00673 HG00733 HG00735 HG00741; do
   TRUTH_VCF=${DIR_TRUTH_VCF}/$SAMPLE.GRCh38_no_alt.deepvariant.vcf.gz
 
-  bash ${PATH_VCF_EVALUATION_SH} $SAMPLE $TRUTH_VCF $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} ${OUTPUT_DIR}/$SAMPLE ${PATH_VCF_PREPROCESS_SH}
+  bash ${PATH_VCF_EVALUATION_SH} $SAMPLE $TRUTH_VCF $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} $SAMPLE ${PATH_VCF_PREPROCESS_SH}
 done
 
 mv ${PREFIX} ${OUTPUT_DIR}

### 2022-02-17 scripts/minihprc_test/mini_hprc.sh
fix dir

@@ -25,18 +25,18 @@ PATH_VCF_EVALUATION_SH=/lizardfs/guarracino/HPRC/mini_dataset/vcf_evaluation.sh
                 zgrep -P "^chr$i\t" ${DIR_REGIONS}/GRCh38_alldifficultregions.bed.gz | bgzip > ${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz;
         fi
 
-        OUTPUT_DIR=${OUTPUT_DIR}/$PREFIX/chr${i}
+        OUTPUT_DIR_CHR=${OUTPUT_DIR}/$PREFIX/chr${i}
 
-        if [ ! -d ${OUTPUT_DIR} ]; then
+        if [ ! -d ${OUTPUT_DIR_CHR} ]; then
                 #echo $PREFIX
 
-                mkdir -p $OUTPUT_DIR
+                mkdir -p $OUTPUT_DIR_CHR
 
                 FASTA=${DIR_FASTA}/chr${i}hg00.fa.gz
                 PATH_EASY_REGIONS=${DIR_REGIONS}/GRCh38_notinalldifficultregions.chr$i.bed.gz
                 PATH_HARD_REGIONS=${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz
 
-                sbatch -p workers -c 48 --job-name mini$PREFIX --wrap 'bash single_hprc_chr.sh '$FASTA' '$PREFIX' '${PATH_WFMASH}' '$OUTPUT_DIR' '${PATH_EASY_REGIONS}' '${PATH_HARD_REGIONS}' '${PATH_FASTA_PAF_TO_VCF}' '${DIR_TRUTH_VCF}' '${PATH_VCF_PREPROCESS_SH}' '${PATH_VCF_EVALUATION_SH}';'
+                sbatch -p workers -c 48 --job-name mini$PREFIX --wrap 'bash single_hprc_chr.sh '$FASTA' '$PREFIX' '${PATH_WFMASH}' '${OUTPUT_DIR_CHR}' '${PATH_EASY_REGIONS}' '${PATH_HARD_REGIONS}' '${PATH_FASTA_PAF_TO_VCF}' '${DIR_TRUTH_VCF}' '${PATH_VCF_PREPROCESS_SH}' '${PATH_VCF_EVALUATION_SH}';'
         fi
 done
 

### 2022-02-17 scripts/minihprc_test/single_hprc_chr.sh

@@ -23,4 +23,4 @@ for SAMPLE in HG00438 HG00621 HG00673 HG00733 HG00735 HG00741; do
   bash ${PATH_VCF_EVALUATION_SH} $SAMPLE $TRUTH_VCF $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} $SAMPLE ${PATH_VCF_PREPROCESS_SH}
 done
 
-mv ${PREFIX} ${OUTPUT_DIR}
+mv ${PREFIX}* ${OUTPUT_DIR}

### 2022-02-19 manuscript/main.tex
update references and main structure

@@
 
-\usepackage[draft]{hyperref}
 \usepackage{makecell}
+
+\usepackage{floatrow}
+
 \usepackage{comment}
 
-\usepackage{floatrow}
+\usepackage{siunitx}
 
@@
 
+\usepackage[usenames,dvipsnames]{xcolor}
+
+\usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
+
 \usepackage{algorithm2e}
-\usepackage[usenames,dvipsnames]{xcolor}
+\SetKwRepeat{Do}{do}{while}%
 
@@
 
+\usepackage{orcidlink}
+\hypersetup{hidelinks}
+
 \def\labelitemi{--}
@@
 \access{Advance Access Publication Date: Day Month Year}
-\appnotes{Sequence Analysis}
+\appnotes{Genome Analysis}
 
@@
 
-\subtitle{Sequence Analysis}
+\subtitle{Genome Analysis}
 
@@
 \author[Guarracino \textit{et~al}.]{
-Andrea~Guarracino\,$^{\text{\sfb 1}}$,
+Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}\text{\sfb 1}}$
 Njagi~Mwaniki\,$^{\text{\sfb 2}}$,
-Santiago~ Marco-Sola\,$^{\text{\sfb 3, 4}}$,
-and~Erik~Garrison\,$^{\text{\sfb 5}*}$
+Santiago~ Marco-Sola\,$^{\orcidlink{0000-0001-7951-3914}\text{\sfb 3, 4}}$,
+Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 5}*}$
 }
@@
 \editor{Associate Editor: XXXXXXX}
-
-% XXX key message of the paper is that we have collected a set of algorithms that enable easy use of pangenome graphs for investigating biology
 
@@
 \section{Introduction}
-Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{22388286, BWA_MEM, 23103880}
-to variant detection~\citep{21478889}, as well as de~novo assembly~\citep{19251739} and pangenome construction~\citep{33177663, 33066802}.
+\label{sec:introduction}
+Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{Langmead2012, li2013aligning, MarcoSola2012} to variant detection~\citep{DePristo2011}, as well as de~novo assembly~\citep{Simpson2009} and pangenome construction~\citep{Armstrong2020, Li2020, pggb-gh}.
 Pairwise sequence alignment can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary relationships between two biological sequences.
-In this context, pangenomes model the full set of genomic elements in a given species or clade~\citep{32453966}.
+In this context, pangenomes model the full set of genomic elements in a given species or clade~\citep{Eizenga_2020}.
 These data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences to a particular genome chosen as reference.
@@
 
-Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory when generating
-sensitive alignments for chromosome-scale contigs available thanks to the new sequencing technologies~\citep{33288905}.
-To scale to vertebrate genomes, methods for pangenome construction based on these approaches must first filter out centromeric
-and other highly-repetitive sequences, in contrast to the pangenome idea of modeling the full genetic variation in the samples.
+Popular seed-and-extend mapping methods, like Minimap2~\citep{Li_2018}, poorly scale in runtime and memory when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing technologies~\citep{Garg_2020}.
+To scale to vertebrate genomes, methods for pangenome construction based on these approaches must first filter out centromeric and other highly-repetitive sequences, in contrast to the pangenome idea of modeling the full genetic variation in the samples.
 
-In the interest of operating on the full pangenome, and to face the upcoming challenges in pangenome construction,
-we have developed wfmash, a new gap-affine pairwise tool for aligning DNA sequences at a whole-chromosome scale.
+In the interest of operating on the full pangenome, and to face the upcoming challenges in pangenome construction, we have developed wfmash, a new gap-affine pairwise tool for aligning DNA sequences at a whole-chromosome scale.
 It supports split-read alignment and gap-affine penalties for insertion and deletions.
-wfmash scales efficiently to large eukaryotic chromosomes while requiring little time and memory, making it suitable to
-be applied on the pairwise alignment of hundreds of genomes.
-wfmash is a DNA sequence aligner based on the integration of MashMap~\citep{30423094}, a fast approximate sequence mapper
-for computing local alignment boundaries between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952},
-an exact gap-affine algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
+wfmash scales efficiently to large eukaryotic chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise alignment of hundreds of genomes.
+wfmash is a DNA sequence aligner based on the integration of MashMap~\citep{Jain_2018}, a fast approximate sequence mapper for computing local alignment boundaries between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{Marco_Sola_2020}, an exact gap-affine algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
 
 \section{Methods}
+\label{sec:methods}
 
-wfmash first applies a locality-sensitive hashing, from MashMap, to rapidly determine syntenic region
-boundaries between long DNA sequences. Then, a hierarchical implementation of the WFA allows computing the
-base-level global alignment of the identified mappings.
+wfmash first applies a locality-sensitive hashing, from MashMap, to rapidly determine syntenic region boundaries between long DNA sequences.
+Then, a hierarchical implementation of the WFA allows computing the base-level global alignment of the identified mappings.
 
-\subsection{Approximate mapping}
+\subsection{Approximate mapping and world minimizers}
 Each query sequence is broken into non-overlapping pieces of the requested length.
-These segments are then mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}.
-We extended MashMap, incorporating the robust winnowing~\citep{Schleimer S. et al. (2003)} in the minimizer downsampling.
-Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in runtime and memory-usage without affecting accuracy~\citep{32657365}.
-Futhermore, we made it possible to set the number of mappings to return for each segment; this is useful for identifying
-paralogous regions and homologous relationship between sets of genomes.
-Together with the segment length and the minimum segment estimated identity, these settings allow users to precisely
-define the mapping space to consider, specifying the characteristics of homologies to compute.
+These segments are then mapped using MashMap's sliding MinHash mapping algorithm~\citep{Jain_2018}.
+We extended MashMap, incorporating the robust winnowing~\citep{Schleimer2003} in the minimizer downsampling.
+Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in runtime and memory-usage without affecting accuracy~\citep{Jain_2020}.
+Furthermore, we made it possible to set the number of mappings to return for each segment; this is useful for identifying paralogous regions and homologous relationship between sets of genomes.
+Together with the segment length and the minimum segment estimated identity, these settings allow users to precisely define the mapping space to consider, specifying the characteristics of homologies to compute.
 \\
@@
 
-\subsection{Base-level alignment}
+\subsection{Base-level alignment with wflambda}
 The time and memory required to compute the base-level alignment increases quadratically with the sequence length.
-The WFA provides an efficient way to decrease the amount of computation required to obtain the optimal alignment between two sequences,
-reducing the cost to be quadratic in the alignment penalty score of the optimal alignment~\citep{32915952}.
-This means that the algorithm is very efficient in aligning similar homologous sequences (i.e., sequences with a
-low alignment penalty score), but high divergence genomes and/or noisy long reads can increase its memory and runtime costs.
+The WFA provides an efficient way to decrease the amount of computation required to obtain the optimal alignment between two sequences, reducing the cost to be quadratic in the alignment penalty score of the optimal alignment~\citep{Marco_Sola_2020}.
+This means that the algorithm is very efficient in aligning similar homologous sequences (i.e., sequences with a low alignment penalty score), but high divergence genomes and/or noisy long reads can increase its memory and runtime costs.
 
@@
 
-The hierarchical implementation requires only the memory to align the sequences at \textit{W}-bps resolution,
-limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
-This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
-no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
+The hierarchical implementation requires only the memory to align the sequences at \textit{W}-bps resolution, limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
+This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
 %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
@@
 
+\subsection{Sparsificatin}
+XXXXXXXXXXXXXXXXXXXXXXX
+
+\subsection{Patching}
+XXXXXXXXXXXXXXXXXXXXXXX
+
 \section{Results}
+\label{sec:results}
 
-We evaluated the performance of wfmash \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{29750242} and
-Winnomap 2.03~\citep{XXX}, in terms of runtime and memory usage. The comparison was performed on both simulated
-and real datasets. Simulated data allowed evaluating the alignment quality by calculating precision and recall
-in variant identification. The evaluations were performed on a workstation running Ubuntu 20.04 equipped with
-64 GB of RAM and an Intel i9-9900K CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.
+\subsection{Input and output features}
+XXXXXXXXXXXXXXXXXXXXXXX
+
+\subsection{liftover completeness experiments}
+XXXXXXXXXXXXXXXXXXXXXXX
+
+\subsection{Pangenome building and evaluation}
+%wfmash/minimap2/lastz/winnomap2 | seqwish | deconstruct | vcfeval
+XXXXXXXXXXXXXXXXXXXXXXX
+
+\subsection{Pangenome alignment (ONT vs HPRC)}
+XXXXXXXXXXXXXXXXXXXXXXX
+
+\begin{comment}
+We evaluated the performance of wfmash \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{Li_2018} and Winnomap 2.03~\citep{Jain_2020b}, in terms of runtime and memory usage.
+The comparison was performed on both simulated and real datasets. Simulated data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
+The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K CPU with 8 cores (16 threads).
+All commands are provided in the Supplementary Material.
 
 \subsection{Simulated data}
-We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and the
-telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of length and divergence
-(\FIXME{MAYBETable 1 and Table 2}). Chromosomes were split into shorter sequences using the splitfa tool (6a0bb67)~\citep{splitfa}.
-In each run, all sequences have the same length, with 50\% of them in reverse complement respect to the source
-chromosome. Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced
-using the Mutation-Simulator script~\citep{32780803}.
+We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of length and divergence (\FIXME{MAYBETable 1 and Table 2}).
+Chromosomes were split into shorter sequences using the splitfa tool (random\_segment\_length branch, 6a0bb67)~\citep{splitfa-gh}.
+In each run, all sequences have the same length, with 50\% of them in reverse complement respect to the source chromosome.
+Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced using the Mutation-Simulator script~\citep{Khl2020}.
 %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.
 
-To evaluate the correctness of the alignments, in each run we used the pairwise alignment of the input sequences to build
-a pangenome graph and identify the variants embedded in it. The graphs were built using~\citep{seqwish}.
-seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
-sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
-evaluating the false-negative and false-positive rates using vcfeval~\citep{vcfeval} for SNVs and small indels,
-and FIXME XXXXX for SVs.
+To evaluate the correctness of the alignments, in each run we used the pairwise alignment of the input sequences to build a pangenome graph and identify the variants embedded in it.
+The graphs were built using~\citep{Garrison2022}.
+seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the sequences and their alignments.
+We called SNVs, small indels, and SVs using vg deconstruct~\citep{Garrison:2018}, evaluating the false-negative and false-positive rates using vcfeval~\citep{Cleary2015} for SNVs and small indels, and FIXME XXXXX for SVs.
 
-\begin{comment}
 \begin{table}[!t]
@@
     }
-
 \end{table}
-\end{comment}
-
-% TODO Simulation with full chromosomes too
-
-% TODO 7 Mbp beta-defensin locus (chr8: 6,300,000-13,300,000): wfmash/minimap2/winnomnap2
-
-% TODO Centromere (chr8:42,881,543-47,029,467): only wfmash
 
 \subsection{Real data}
-XXXXX
 \\
@@
 
+\end{comment}
+
 \section{Discussion}
-We implemented a novel gap-affine pairwise aligner, wfmash, to accelerate the computation of the mutual
-alignment of collections of genomes, a step required for constructing pangenome models. We have demonstrated that
-it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
-Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, wfmash is already
-successfully applied in our pangenome graphs building pipeline~\citep{pggb}. No less important, pairwise
-alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the
-increasing yields of sequencing technologies in the coming years.
+\label{sec:discussion}
+
+We implemented a novel gap-affine pairwise aligner, wfmash, to accelerate the computation of the mutual alignment of collections of genomes, a step required for constructing pangenome models.
+We have demonstrated that it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the Human Pangenome Reference Consortium year 1 assembly.
+Indeed, thanks to its efficiency, wfmash is already successfully applied in our pangenome graphs building pipeline~\citep{pggb-gh}.
+No less important, pairwise alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the increasing yields of sequencing technologies in the coming years.
+
+\subsection{Mashmap limits and potential improvements}
+
+XXXXX
 
@@
 We thank members of the HPRC Pangenome Working Group for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
+%todo VGP/EBP too?
 
@@
 
-We gratefully acknowledge support from NIH/NIDA U01DA047638 and NIH/NIGMS R01GM123489 (EG).
+We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG) and efforts by Dr. Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy (AG).
+\linebreak
+\linebreak
+\textit{Conflict of Interest}: none declared.
 
@@
 \bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%
-%\bibliographystyle{plain}
-%
+
 \bibliography{document}
 
-% \begin{thebibliography}{}
-
-% \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-% Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
-
-% \bibitem[Bag {\it et~al}., 2001]{Bag01}
-% Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
-
-% \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-% Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-% in nigral dopaminergic neurnol cell: correlation with the known
-% pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-% Res.}, \textbf{110}(Suppl. 1), 76--84.
-
-% \bibitem[Lehmann, 1986]{Leh86}
-% Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
-
-% \bibitem[Crenshaw and Jones, 2003]{Cre03}
-% Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-% cancer management: one tumor, one chip. \textit{Bioinformatics},
-% doi:10.1093/bioinformatics/btn000.
-
-% \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-% Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-% (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-% ???--???.
-
-% \bibitem[Bardet, 1920]{Bar20}
-% Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-% polydactylie et retinite pigmentaire (contribution a l'etude des
-% formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-% institution, Paris, France.
-
-% \end{thebibliography}
 \end{document}

### 2022-02-19 scripts/mapq_evaluation.sh

@@ -1,0 +1,69 @@
+# Download the T2T assembly
+wget -c https://s3-us-west-2.amazonaws.com/human-pangenomics/T2T/CHM13/assemblies/chm13.draft_v1.1.fasta.gz
+
+gunzip chm13.draft_v1.1.fasta.gz
+samtools faidx chm13.draft_v1.1.fasta
+samtools faidx chm13.draft_v1.1.fasta $(grep chr22 chm13.draft_v1.1.fasta.fai | cut -f 1) > chr22.fa && samtools faidx chr22.fa
+
+run_splitfa=~/git/fork/splitfa/target/release/splitfa # https://github.com/AndreaGuarracino/splitfa.git, random_segment_length branch
+
+
+threads=16
+fasta=chr22.fa
+fasta_name=$(basename $fasta .fa)
+
+
+# Create simulated reads
+for len in 200 500 1000 2000 5000 10000 20000 50000 100000 500000 1000000; do
+  ~/git/PBSIM-PacBio-Simulator/src/pbsim --depth 2 --data-type CLR --accuracy-mean 0.99 --accuracy-min 0.99 --length-min $len --length-mean $len --length-max $len --model_qc ~/git/PBSIM-PacBio-Simulator/data/model_qc_clr $fasta
+  ~/git/minimap2/misc/paftools.js pbsim2fq $fasta sd_0001.maf | sed 's/!>/!/g' | bgzip -@ $threads -c > pbsim.${fasta_name}.${len}bp.fa.gz && samtools faidx pbsim.${fasta_name}.${len}bp.fa.gz
+  rm sd_0001.*
+done
+
+# Mapping and alignment
+for len in 200 500 1000 2000 5000 10000 20000 50000 100000 500000 1000000; do
+  s=$(echo "$len/10" | bc)
+  s=$(( $s > 100 ? $s : 100 ))
+  s=$(( $s < 50000 ? $s : 50000 ))
+
+  \time -v minimap2 $fasta pbsim.${fasta_name}.${len}bp.fa.gz -t $threads -c -x asm5 > pbsim.${fasta_name}.${len}bp.mp2.paf
+done
+
+for len in 200 500 1000 2000 5000 10000 20000 50000 100000 500000 1000000; do
+  s=$(echo "$len/10" | bc)
+  s=$(( $s > 100 ? $s : 100 ))
[…]

### 2022-02-19 manuscript/fig_wflamba.tex
fill the structure

\begin{figure}[ht!]
    \begin{subfigure}{\linewidth}
        %\caption{}
        \centering
        % include first image
        \includegraphics[width=1.0\linewidth, trim=-0cm 2cm 0 0cm]{fig/wflambda/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_bad_sorted}
        \label{fig:bad-sorting}
    \end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
    \caption{
        PLACEHOLDER IMAGE XXX XXX XXX
    }
    \label{fig:wflambda}
\end{figure}

### 2022-02-19 manuscript/main.tex
fill the structure

@@
 
-\title[wfmash: whole-cromosome pairwise aligner]{wfmash: whole-chromosome pairwise alignment using the hierarchical wavefront algorithm}
+\title[Whole-cromosome pairwise alignment]{Whole-chromosome pairwise alignment using the hierarchical wavefront algorithm}
 \author[Guarracino \textit{et~al}.]{
@@
 \textbf{Motivation:}
-Pairwise alignment of sequences is an important step in many bioinformatics analyses, including pangenome construction.
-Pangenomes are sequence models able to provide a full representation of the mutual alignment of collections of genomes.
-The time and memory required to compute pairwise alignments increase quadratically with the sequence length,
-making it impractical to directly align very long sequences without applying heuristic approaches to first determine possible syntenic regions in the sequences.
-Nevertheless, with the advances in sequencing technologies, new genome assemblies are produced at a high rate,
-pressing for the development of tools able to align sequences of the order of tens of megabases long. \\
+Pairwise alignment is a critical step in studying whole-genome assemblies and their relationships in an unbiased way.
+The time and memory required to compute the pairwise alignments increase quadratically with the sequence length.
+This makes it impractical to align very long sequences without applying heuristic approaches to first determine possible synthenic regions in the sequences.
+Nevertheless, with the advances in sequencing technologies, new whole-genome assemblies are produced at a high rate, pressing for the development of tools able to align sequences of the order of tens of megabases long.
+\\
 \textbf{Results:}
@@
 \label{sec:introduction}
-Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{Langmead2012, li2013aligning, MarcoSola2012} to variant detection~\citep{DePristo2011}, as well as de~novo assembly~\citep{Simpson2009} and pangenome construction~\citep{Armstrong2020, Li2020, pggb-gh}.
+Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{Langmead2012, li2013aligning, MarcoSola2012} to variant detection~\citep{DePristo2011}, as well as de~novo assembly~\citep{Simpson2009} and pangenome construction~\citep{Armstrong2020, Li2020, pggb}.
 Pairwise sequence alignment can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary relationships between two biological sequences.
@@
 The unbiased approach allows studying the entire genetic diversity of a population.
-This opportunity, in combination with the massive amount of data available nowadays, presses for the development of tools able to compute pairwise alignment of very long sequences.
+This opportunity, in combination with the massive amount of haplotype-resolved whole-genome assemblies available nowadays~\citep{Garg_2020}, presses for the development of tools able to compute pairwise alignments of very long sequences.
 
-Popular seed-and-extend mapping methods, like Minimap2~\citep{Li_2018}, poorly scale in runtime and memory when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing technologies~\citep{Garg_2020}.
-To scale to vertebrate genomes, methods for pangenome construction based on these approaches must first filter out centromeric and other highly-repetitive sequences, in contrast to the pangenome idea of modeling the full genetic variation in the samples.
+Popular seed-and-extend mapping methods, like Minimap2~\citep{Li_2018}, poorly scale in runtime and memory when generating sensitive alignments for chromosome-scale contigs.
+To scale up to vertebrate genomes, pangenome construction pipelines based on these methods must first filter out centromeres and other highly-repetitive sequences that are typical in eukaryotic chromosomes.
+However, removing sequences is in contrast to the pangenome idea of modeling the full genetic variation in the samples.
+%TODO SOMETHING ELSE REGARDING OTHER ALIGNMENT METHODS???????????????????????????????????????????????
 
@@
 It supports split-read alignment and gap-affine penalties for insertion and deletions.
-wfmash scales efficiently to large eukaryotic chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise alignment of hundreds of genomes.
-wfmash is a DNA sequence aligner based on the integration of MashMap~\citep{Jain_2018}, a fast approximate sequence mapper for computing local alignment boundaries between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{Marco_Sola_2020}, an exact gap-affine algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
+wfmash scales efficiently to large chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise alignment of hundreds of genomes.
+wfmash is based on the integration of MashMap~\citep{Jain_2018}, a fast approximate sequence mapper for computing local alignment boundaries between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{Marco_Sola_2020}, an exact gap-affine algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
 
@@
 
-\subsection{Approximate mapping and world minimizers}
+\subsection{Approximate mapping}
 Each query sequence is broken into non-overlapping pieces of the requested length.
 These segments are then mapped using MashMap's sliding MinHash mapping algorithm~\citep{Jain_2018}.
-We extended MashMap, incorporating the robust winnowing~\citep{Schleimer2003} in the minimizer downsampling.
-Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in runtime and memory-usage without affecting accuracy~\citep{Jain_2020}.
+We extended MashMap by incorporating the robust winnowing~\citep{Schleimer2003} in the minimizer sampling.
+Robust sampling avoids taking too many minimizers in low-complexity regions in the sequences, yielding improvements in runtime and memory-usage without affecting accuracy~\citep{Jain_2020}.
 Furthermore, we made it possible to set the number of mappings to return for each segment; this is useful for identifying paralogous regions and homologous relationship between sets of genomes.
 Together with the segment length and the minimum segment estimated identity, these settings allow users to precisely define the mapping space to consider, specifying the characteristics of homologies to compute.
-\\
-FIXME THINGS ON THE FILTERING???
-\\
-FIXME THING ON THE SPLIT READ AND/OR THE MERGING???
-\\
+%TODO THINGS ON THE FILTERING???
+%TODO THING ON THE SPLIT READ AND/OR THE MERGING???
 
-Finally, each mapping location is used as a target for base-level alignment using a hierarchical implementation
-of the WFA.
+\FIXME{World minimizers XXXXXX}
+
+\FIXME{Spaced-seed XXXXXX}
 
 \subsection{Base-level alignment with wflambda}
-The time and memory required to compute the base-level alignment increases quadratically with the sequence length.
-The WFA provides an efficient way to decrease the amount of computation required to obtain the optimal alignment between two sequences, reducing the cost to be quadratic in the alignment penalty score of the optimal alignment~\citep{Marco_Sola_2020}.
-This means that the algorithm is very efficient in aligning similar homologous sequences (i.e., sequences with a low alignment penalty score), but high divergence genomes and/or noisy long reads can increase its memory and runtime costs.
+Each mapping location is used as a target for base-level alignment using a hierarchical implementation of the WFA.
+Time and space computational complexities of a base-level alignment are quadratic with respect to the sequence length.
+The WFA provides an efficient way to decrease the amount of computation required to obtain the optimal alignment between two sequences, reducing the computational cost to be quadratic in the alignment penalty score of the optimal alignment~\citep{Marco_Sola_2020}.
+This means that WFA is very efficient in aligning similar homologous sequences (i.e., sequences with a low alignment penalty score), but high divergence genomes and/or noisy long reads can increase its time and memory costs.
 
 To avoid such limitation, wfmash applies a hierarchical WFA, exploiting the WFA to guide the alignment process, but keeping the largest alignment problem size small.
-Rather than directly aligning the whole sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment (Figure~1\vphantom{\ref{fig:1}}).
-The whole global alignment is then computed over the full dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
-Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments \textit{W}-bps long from the two sequences to align.
-To determine if each cell is a match, and then guiding the \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the two fragments.
+Rather than directly aligning the whole sequences, it aligns them to each other in small pieces \textit{W}-bp long by applying a global WFA alignment (Figure~1\vphantom{\ref{fig:1}}).
+The whole global alignment is then computed over the full dynamic-programming matrix (high order DP-matrix) at \textit{W}-bp resolution.
+Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments \textit{W}-bp long from the two sequences to align.
+
+\input{fig_wflamba}
+
+\subsection{Sparsification}
+To determine if each cell is a match, and then guiding the alignment at \textit{W}-bp resolution, the global alignment is performed between the two fragments with the standard WFA.
 To accelerate the process, an alignment-free comparison of the two fragments is performed first by computing the mash distance between them.
@@
 
-Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments on the optimal path in the high order DP-matrix.
+\subsection{Patching}
+Finally, the \textit{W}-bp resolution traceback is applied to determine the set of base-level alignments on the optimal path in the high order DP-matrix.
 During the process, a further step is performed for refining the alignment.
@@
 
-The hierarchical implementation requires only the memory to align the sequences at \textit{W}-bps resolution, limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
-This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
+\subsection{Complexity}
+The hierarchical implementation requires only the memory to align the sequences at \textit{W}-bp resolution, limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bp long.
+This approach is more \FIXME{flexible than using a fixed-width band (it effectively has a variable bandwidth), requires no heuristic seeding step}, and can benefit from parallel exploration of the wavefront.
 %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
-%TODO 'and can benefit from parallel exploration of the wavefront.' Can we?
-
-\input{fig_sorting}
-
-\subsection{Sparsificatin}
-XXXXXXXXXXXXXXXXXXXXXXX
-
-\subsection{Patching}
-XXXXXXXXXXXXXXXXXXXXXXX
+\\
 
@@
 \subsection{Input and output features}
-XXXXXXXXXXXXXXXXXXXXXXX
+XXXXXXXXXXXXX -s, -l, -p, -n, -i, sam/paf
 
-\subsection{liftover completeness experiments}
+\subsection{Liftover completeness}
 XXXXXXXXXXXXXXXXXXXXXXX
@@
 \subsection{Pangenome building and evaluation}
-%wfmash/minimap2/lastz/winnomap2 | seqwish | deconstruct | vcfeval
-XXXXXXXXXXXXXXXXXXXXXXX
+We evaluated the performance of wfmash \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{Li_2018} and Winnomap 2.03~\citep{Jain_2020b}, in terms of runtime and memory usage.
+The comparison was performed on both simulated and real datasets. Simulated data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
+The evaluations were performed on a workstation running \FIXME{Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K CPU with 8 cores (16 threads)}.
+
+We simulated long sequences from \FIXME{the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and the telomere-to-telomere (T2T) assemblies of the human chromosomes 8}, at different levels of length and divergence (\FIXME{MAYBETable 1 and Table 2}).
+Chromosomes were split into shorter sequences using the splitfa tool (6a0bb67)~\citep{splitfa-gh}.
+In each run, all sequences have the same length, with half of them in reverse complement respect to the source chromosome.
+Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced using the Mutation-Simulator script~\citep{Khl2020}.
+\FIXME{Structural variants (SVs) were applied using the SURVIVOR benchmarking tool}~\citep{Jeffares2017}.
+
+To evaluate the correctness of the alignments, we used the pairwise alignment of the input sequences from each run to build a pangenome graph and identified the variants embedded in it.
+The graphs were built with seqwish~\citep{Garrison2022}, a lossless pangenome graph inducer.
+We called SNVs, small indels, and SVs using vg deconstruct~\citep{Garrison:2018}, evaluating the false-negative and false-positive rates using vcfeval~\citep{Cleary2015} for SNVs and small indels, and \FIXME{XXXXX for SVs}.
 
 \subsection{Pangenome alignment (ONT vs HPRC)}
-XXXXXXXXXXXXXXXXXXXXXXX
+XXXXXXXXXXXXX
 
 \begin{comment}
-We evaluated the performance of wfmash \texttt{90ea671} against Minimap2-v2.20-r1061~\citep{Li_2018} and Winnomap 2.03~\citep{Jain_2020b}, in terms of runtime and memory usage.
-The comparison was performed on both simulated and real datasets. Simulated data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
-The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K CPU with 8 cores (16 threads).
-All commands are provided in the Supplementary Material.
-
-\subsection{Simulated data}
-We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of length and divergence (\FIXME{MAYBETable 1 and Table 2}).
-Chromosomes were split into shorter sequences using the splitfa tool (random\_segment\_length branch, 6a0bb67)~\citep{splitfa-gh}.
-In each run, all sequences have the same length, with 50\% of them in reverse complement respect to the source chromosome.
-Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced using the Mutation-Simulator script~\citep{Khl2020}.
-%, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.
-
-To evaluate the correctness of the alignments, in each run we used the pairwise alignment of the input sequences to build a pangenome graph and identify the variants embedded in it.
-The graphs were built using~\citep{Garrison2022}.
-seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the sequences and their alignments.
-We called SNVs, small indels, and SVs using vg deconstruct~\citep{Garrison:2018}, evaluating the false-negative and false-positive rates using vcfeval~\citep{Cleary2015} for SNVs and small indels, and FIXME XXXXX for SVs.
 
@@
 We have demonstrated that it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the Human Pangenome Reference Consortium year 1 assembly.
-Indeed, thanks to its efficiency, wfmash is already successfully applied in our pangenome graphs building pipeline~\citep{pggb-gh}.
+Indeed, thanks to its efficiency, wfmash is already successfully applied in our pangenome graphs building pipeline~\citep{pggb}.
 No less important, pairwise alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the increasing yields of sequencing technologies in the coming years.
 
-\subsection{Mashmap limits and potential improvements}
+\subsection{Limits and potential improvements}
 
@@
 
-%Data used to build human pangenome graphs: \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}. 
-%The graphs we generated the figures from: \url{https://s3-us-west-2.amazonaws.com/human-pangenomics/index.html?prefix=pangenomes/scratch/2021_11_04_pggb_wgg.87/}.
-Code and links to data resources used to build this manuscript and its figures, can be found in the paper's public repository: \url{https://github.com/AndreaGuarracino/wfmash-paper}.
+Code and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/AndreaGuarracino/wfmash-paper}.
 

### 2022-02-24 scripts/minihprc_test/fasta+paf2vcf.mod.sh
avoid file collisions

@@ -25,11 +25,11 @@ GFA=$PAF.k0.B10M.gfa
 VCF=$GFA.vcf.gz
 
 echo "All-vs-all alignment"
-\time -v $PATH_WFMASH $FASTA $FASTA -X -s ${SEGMENT_LEN} -l ${BLOCK_LEN} -p $IDENTITY -n ${NUM_HAPLOTYPES} -k 16 -t 48 > $PAF
+\time -v $PATH_WFMASH $FASTA $FASTA -X -s ${SEGMENT_LEN} -l ${BLOCK_LEN} -p $IDENTITY -n ${NUM_HAPLOTYPES} -k 16 -t 12 > $PAF
 
 echo "Graph induction"
 # -k 0` is for getting a lossless representation of the pairwise alignment
-\time -v ${PATH_SEQWISH} -s $FASTA -p $PAF -g $GFA -k 0 -B 10M -t 48 -P
+\time -v ${PATH_SEQWISH} -s $FASTA -p $PAF -g $GFA -k 0 -B 10M -t 12 -P
 
 echo "Identify variants"
-\time -v ${PATH_VG} deconstruct -e -a -P $REFERENCE_PREFIX -H $SEPARATOR $GFA -t 48 | bgzip -c > $VCF && tabix $VCF
+\time -v ${PATH_VG} deconstruct -e -a -P $REFERENCE_PREFIX -H $SEPARATOR $GFA -t 12 | bgzip -c > $VCF && tabix $VCF

### 2022-02-24 scripts/minihprc_test/mini_hprc.sh

@@ -36,7 +36,7 @@ PATH_VCF_EVALUATION_SH=/lizardfs/guarracino/HPRC/mini_dataset/vcf_evaluation.sh
                 PATH_EASY_REGIONS=${DIR_REGIONS}/GRCh38_notinalldifficultregions.chr$i.bed.gz
                 PATH_HARD_REGIONS=${DIR_REGIONS}/GRCh38_alldifficultregions.chr$i.bed.gz
 
-                sbatch -p workers -c 48 --job-name minihprc --wrap 'bash single_hprc_chr.sh '$FASTA' '$PREFIX' '${PATH_WFMASH}' '${OUTPUT_DIR_CHR}' '${PATH_EASY_REGIONS}' '${PATH_HARD_REGIONS}' '${PATH_FASTA_PAF_TO_VCF}' '${DIR_TRUTH_VCF}' '${PATH_VCF_PREPROCESS_SH}' '${PATH_VCF_EVALUATION_SH}';'
+                sbatch -p workers -c 12 --job-name minihprc --wrap 'bash single_hprc_chr.sh '$i' '$FASTA' '$PREFIX' '${PATH_WFMASH}' '${OUTPUT_DIR_CHR}' '${PATH_EASY_REGIONS}' '${PATH_HARD_REGIONS}' '${PATH_FASTA_PAF_TO_VCF}' '${DIR_TRUTH_VCF}' '${PATH_VCF_PREPROCESS_SH}' '${PATH_VCF_EVALUATION_SH}';'
         fi
 done
 

### 2022-02-24 scripts/minihprc_test/single_hprc_chr.sh

@@ -1,18 +1,19 @@
 #!/bin/bash
 
-FASTA=$1
-PREFIX=$2
-PATH_WFMASH=$3
-OUTPUT_DIR=$4
-PATH_EASY_REGIONS=$5
-PATH_HARD_REGIONS=$6
-PATH_FASTA_PAF_TO_VCF=$7
-DIR_TRUTH_VCF=$8
-PATH_VCF_PREPROCESS_SH=$9
-PATH_VCF_EVALUATION_SH=${10}
+i=$1
+FASTA=$2
+PREFIX=$3
+PATH_WFMASH=$4
+OUTPUT_DIR=$5
+PATH_EASY_REGIONS=$6
+PATH_HARD_REGIONS=$7
+PATH_FASTA_PAF_TO_VCF=$8
+DIR_TRUTH_VCF=$9
+PATH_VCF_PREPROCESS_SH=${10}
+PATH_VCF_EVALUATION_SH=${11}
 
-mkdir -p /scratch/$PREFIX
-cd /scratch/$PREFIX
+mkdir -p /scratch/$PREFIX-chr$i
+cd /scratch/$PREFIX-chr$i
 
 bash ${PATH_FASTA_PAF_TO_VCF} $FASTA 100k 300k 98 14 grch38 '#' ${PATH_WFMASH} ${PREFIX};
 
@@ -25,5 +26,5 @@ for SAMPLE in HG00438 HG00621 HG00673 HG00733 HG00735 HG00741; do
   bash ${PATH_VCF_EVALUATION_SH} $SAMPLE $TRUTH_VCF $VCF ${PATH_EASY_REGIONS} ${PATH_HARD_REGIONS} $SAMPLE ${PATH_VCF_PREPROCESS_SH}
 done
 
-mv /scratch/$PREFIX/* ${OUTPUT_DIR}
-rm /scratch/$PREFIX/ -rf
+mv /scratch/$PREFIX-chr$i/* ${OUTPUT_DIR}
+rm /scratch/$PREFIX-chr$i/ -rf

### 2022-02-24 scripts/minihprc_test/vcf_evaluation.sh

@@ -14,7 +14,7 @@ OUTPUT_DIR=$6
 PATH_VCF_PREPROCESS_SH=$7
 
 echo "VCF renaming"
-zcat $QUERY_VCF | sed 's/^grch38#//g' | bgzip -c -@ 48 > "$QUERY_VCF".renamed.vcf.gz && tabix "$QUERY_VCF".renamed.vcf.gz
+zcat $QUERY_VCF | sed 's/^grch38#//g' | bgzip -c -@ 12 > "$QUERY_VCF".renamed.vcf.gz && tabix "$QUERY_VCF".renamed.vcf.gz
 
 echo "VCF processing"
 bash ${PATH_VCF_PREPROCESS_SH} "$QUERY_VCF".renamed.vcf.gz "$SAMPLE" 50
@@ -31,7 +31,7 @@ echo "VCF evaluation"
     -b "$TRUTH_VCF" \
     -e "$EASY_REGIONS_BED" \
     -c "$NORMALIZED_VCF" \
-    -T 48 \
+    -T 12 \
     -o "$OUTPUT_DIR"/easy
 
 /gnu/store/3vmp4dw8y0r49h0hbjbgv3bckgvz4k0m-rtg-tools-3.11/rtg vcfeval \
@@ -39,5 +39,5 @@ echo "VCF evaluation"
     -b "$TRUTH_VCF" \
     -e "$HARD_REGIONS_BED" \
     -c "$NORMALIZED_VCF" \
-    -T 48 \
+    -T 12 \
     -o "$OUTPUT_DIR"/hard

### 2022-03-30 manuscript/fig_wflamba.tex
figure

@@
         % include first image
-        \includegraphics[width=1.0\linewidth, trim=-0cm 2cm 0 0cm]{fig/wflambda/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_bad_sorted}
+        \includegraphics[width=1.0\linewidth, trim=-0cm 2cm 0 0cm]{fig/wflign.png}
         \label{fig:bad-sorting}
@@
     \caption{
-        PLACEHOLDER IMAGE XXX XXX XXX
+      \textit{wflign} exploration of the alignment of $\sim$560kbp homologus segments of two \textit{Escherichia coli} genomes (GenBank NC\_000913.3 and NC\_008563.1).
+      Gray indicates cells explored by WF$\lambda$.
+      Blue cells are attempted, failed local WFA alignments, while green cells indicate alignments that match the maximum local alignment score.
+      %The traceback (not shown) drives through the 
     }

### 2022-03-30 manuscript/main.tex
update simulations
figure

@@
 
-\title[Whole-cromosome pairwise alignment]{Whole-chromosome pairwise alignment using the hierarchical wavefront algorithm}
+\title[Whole-genome pairwise alignment]{Whole-genome pairwise alignment using the hierarchical wavefront algorithm}
 \author[Guarracino \textit{et~al}.]{
-Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}\text{\sfb 1}}$
+Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}\text{\sfb 1}}$,
 Njagi~Mwaniki\,$^{\text{\sfb 2}}$,
@@
 $^\ast$To whom correspondence should be addressed.
-%$^\dagger$Contributed equally.
+$^\dagger$Contributed equally.
 }
@@
 Pairwise alignment is a critical step in studying whole-genome assemblies and their relationships in an unbiased way.
-The time and memory required to compute the pairwise alignments increase quadratically with the sequence length.
+The time and memory required to compute the pairwise alignments increase quadratically with sequence length and nucleotide divergence.
 This makes it impractical to align very long sequences without applying heuristic approaches to first determine possible synthenic regions in the sequences.

### 2022-03-30 manuscript/main.tex
verbose affiliations

@@
 \address{
-$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Milan, Italy \\
-$^{\text{\sf 2}}$Department of Computer Sciences, University of Pisa, Pisa, Italy \\
-$^{\text{\sf 3}}$Department of Computer Sciences, Barcelona Supercomputing Center, Barcelona 08034, Spain \\
-$^{\text{\sf 4}}$Departament d’Arquitectura de Computadors i Sistemes Operatius, Universitat Autònoma de Barcelona, Barcelona 08193, Spain \\
-$^{\text{\sf 5}}$University of Tennessee Health Science Center, Memphis, TN, USA
+$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Viale Rita Levi‑Montalcini 1, Milan, 20157, Italy \\
+$^{\text{\sf 2}}$Department of Computer Sciences, University of Pisa, Pisa, 56126, Italy \\
+$^{\text{\sf 3}}$Department of Computer Sciences, Barcelona Supercomputing Center, Barcelona, 08034, Spain \\
+$^{\text{\sf 4}}$Departament d’Arquitectura de Computadors i Sistemes Operatius, Universitat Autònoma de Barcelona, Barcelona, 08193, Spain \\
+$^{\text{\sf 5}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA
 }
@@
 $^\ast$To whom correspondence should be addressed.
-$^\dagger$Contributed equally.
+%$^\dagger$Contributed equally.
 }

### 2022-03-31 manuscript/main.tex
fix PISA postalcode

--- 2022-03-30 manuscript/main.tex
@@
 \address{
-$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Milan, Italy \\
-$^{\text{\sf 2}}$Department of Computer Sciences, University of Pisa, Pisa, Italy \\
-$^{\text{\sf 3}}$Department of Computer Sciences, Barcelona Supercomputing Center, Barcelona 08034, Spain \\
-$^{\text{\sf 4}}$Departament d’Arquitectura de Computadors i Sistemes Operatius, Universitat Autònoma de Barcelona, Barcelona 08193, Spain \\
-$^{\text{\sf 5}}$University of Tennessee Health Science Center, Memphis, TN, USA
+$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Viale Rita Levi‑Montalcini 1, Milan, 20157, Italy \\
+$^{\text{\sf 2}}$Department of Computer Sciences, University of Pisa, Pisa, 56127, Italy \\
+$^{\text{\sf 3}}$Department of Computer Sciences, Barcelona Supercomputing Center, Barcelona, 08034, Spain \\
+$^{\text{\sf 4}}$Departament d’Arquitectura de Computadors i Sistemes Operatius, Universitat Autònoma de Barcelona, Barcelona, 08193, Spain \\
+$^{\text{\sf 5}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA
 }

### 2022-04-01 manuscript/main.tex
abstract

@@
 \textbf{Motivation:}
-Pairwise alignment is a critical step in studying whole-genome assemblies and their relationships in an unbiased way.
-The time and memory required to compute the pairwise alignments increase quadratically with sequence length and nucleotide divergence.
-This makes it impractical to align very long sequences without applying heuristic approaches to first determine possible synthenic regions in the sequences.
-Nevertheless, with the advances in sequencing technologies, new whole-genome assemblies are produced at a high rate, pressing for the development of tools able to align sequences of the order of tens of megabases long.
+Pairwise alignment is essential to understanding DNA sequence variation.
+The time and memory required to compute pairwise alignments increases quadratically with sequence length and nucleotide divergence.
+Advances in sequencing technology have resulted in the rapid production of new whole-genome sequences.
+This leads to a new problem of efficiently comparing many whole genomes, a costly task for methods designed to align shorter (100bp--100kbp) sequence reads to reference genomes. \\
+%This makes it impractical to align very long sequences without applying heuristic approaches to first determine possible synthenic regions in the sequences.
+%Nevertheless, with the advances in sequencing technologies, new whole-genome assemblies are produced at a high rate, pressing for the development of tools able to align sequences of the order of tens of megabases long.
 \\
 \textbf{Results:}
-Here we present wfmash, a new gap-affine pairwise aligner designed to align DNA sequences at a whole-chromosome scale.
-wfmash applies a hierarchical implementation of the wavefront alignment algorithm to guide the alignment of very long sequences.
-It scales efficiently to large eukaryotic chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little time and memory. \\
+Here we present \textit{wfmash}, a sequence aligner specialized for large collections of very long and highly divergent genome sequences.
+It first finds large syntenic homologies between genomes using locality sensitive hashing, approximate match chaining, and logarithmic-time filtering of these mappings.
+%Two parameters define these mappings: a seed segment length and minimum average nucleotide identity between matched segments, allowing for intuitive experimental setup.
+The resulting mappings can cover entire chromosomes and contain megabase-scale SVs, which presents a significant challenge for direct alignment.
+We address this with an extension of the wavefront algorithm (WFA) that matches segments of sequence rather than single characters.
+Our approach achieves a many thousand-fold reduction in time and space complexity relative to direct alignment with WFA.
+We demonstrate that this comes at little to no cost in terms of variant detection accuracy relative to established sequence aligners. \\
+%\textit{wfmash} allows us to scale to all-to-all alignment of hundreds of vertebrate \\
+%A key feature of \textit{wfmash} is that users can intuitively shape the scope of an alignment experiment using two parameters: a mapping segment length and a minimum average nucleotide identity between segments.
+\\
 \textbf{Availability:}

### 2022-04-01 manuscript/main.tex
introduction

@@
 \label{sec:introduction}
-Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{Langmead2012, li2013aligning, MarcoSola2012} to variant detection~\citep{DePristo2011}, as well as de~novo assembly~\citep{Simpson2009} and pangenome construction~\citep{Armstrong2020, Li2020, pggb}.
+Sequence alignment determines edits that transform one sequence into another.
+This is a core task in biological analysis, with applications ranging from short read~mapping~\citep{Langmead2012, li2013aligning, MarcoSola2012}, variant detection~\citep{DePristo2011}, to de~novo assembly~\citep{Simpson2009} and pangenome construction~\citep{Armstrong2020, Li2020, pggb}.
 Pairwise sequence alignment can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary relationships between two biological sequences.
-In this context, pangenomes model the full set of genomic elements in a given species or clade~\citep{Eizenga_2020}.
-These data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences to a particular genome chosen as reference.
-The unbiased approach allows studying the entire genetic diversity of a population.
-This opportunity, in combination with the massive amount of haplotype-resolved whole-genome assemblies available nowadays~\citep{Garg_2020}, presses for the development of tools able to compute pairwise alignments of very long sequences.
+It is key to our study of evolution.
 
-Popular seed-and-extend mapping methods, like Minimap2~\citep{Li_2018}, poorly scale in runtime and memory when generating sensitive alignments for chromosome-scale contigs.
-To scale up to vertebrate genomes, pangenome construction pipelines based on these methods must first filter out centromeres and other highly-repetitive sequences that are typical in eukaryotic chromosomes.
-However, removing sequences is in contrast to the pangenome idea of modeling the full genetic variation in the samples.
-%TODO SOMETHING ELSE REGARDING OTHER ALIGNMENT METHODS???????????????????????????????????????????????
+Due to the intrinsic quadratic difficulty of exact alignment algorithms, and the concomitant need for heuristic algorithms that render them practical, solutions to the problem of sequence alignment are tied closely to the DNA sequence types being considered.
+However, the basic concepts behind sequence alignment have not changed in more than 30 years.
+Methods like FASTA \citep{Pearson_1988}, BLAST \citep{Altschul_1990}, BLAT \citep{Kent_2002}, and LASTZ \citep{harris2007lastz} focused on high sensitivity alignment driven by short match (e.g.
+$k$-mer) seeds, chaining and filtering of seeds to find candidate alignments, and final base level alignment using algorithms like Smith-Waterman-Gotoh \citep{Gotoh_1982}.
+Short read sequencing led to a deluge of sequencing data that led to a wave of aligners focused on the problem of matching short (100-250bp) reads with few errors to reference genomes of several gigabase-pairs \cite{Langmead2012,MarcoSola2012,li2013aligning}.
+These methods employ various full text indexes to speed up match discovery, including the FM-index \citep{Ferragina_2000}, which supports linear-time exact matching queries that are ideal for application to precise alignment of high-accuracy sequence reads.
 
-In the interest of operating on the full pangenome, and to face the upcoming challenges in pangenome construction, we have developed wfmash, a new gap-affine pairwise tool for aligning DNA sequences at a whole-chromosome scale.
-It supports split-read alignment and gap-affine penalties for insertion and deletions.
-wfmash scales efficiently to large chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise alignment of hundreds of genomes.
-wfmash is based on the integration of MashMap~\citep{Jain_2018}, a fast approximate sequence mapper for computing local alignment boundaries between long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{Marco_Sola_2020}, an exact gap-affine algorithm that takes advantage of homologous regions between the sequences to accelerate the alignment process.
+However, increasing sequence read lengths, and the arrival of whole genome assemblies for many species rendered such approaches more sensitive than necessary to achieve near-perfect alignment performance when mapping long reads to reference genomes.
+Methods like minimap2 \citep{Li_2018}, LRA \citep{Ren_2021}, Winnowmap \citep{Jain_2020}, and MashMap \citep{Jain_2018} compute mappings, or candidate alignments, using a sparse subset of all $k$-mers, which are often referred to as ``minimizers'' due to their definition as hash-minimizing $k$-mers in a given window \cite{Roberts_2004}.
+Rather than considering all $k$-mers, \textit{windowed minimizer} sampling collects $\sim L/w$ $k-mers$ from a sequence of length $L$, wherein each $k$-mer minimizes a given hash function within a window of size $w$.
+This sampling scheme provides guarantees about the distribution of sampled $k$-mers, which aids read alignment and mapping chaining.
+However, it has been discovered that window minimziers are biased and may underestimate the true similarity of sequences \citep{Belbasi_2022}.
+In contrast, \textit{world minimizers}, which sparsify over the entire potential space of $k$-mers \citep{Broder_1997}, are not biased, and have proved useful as the basis for the whole genome comparison system Mash \citep{Ondov_2016}.
+
+With the arrival of many whole genome assemblies for vertebrates \citep{Rhie_2021}, as well as the first complete assembly of the human genome \cite{Nurk_2022}, the length of sequences that we seek to align has increased again by several orders of magnitude.
+Instead of aligning sequences of 100kb, we now must consider the alignment of whole chromosomes of 100Mbp.
+At the same time, the development of pangenomes based on collections of many such genomes (as in the human pangenome project, HPRC) increases the size of a mapping target from 1Gbp to 1Tbp.
+This $10^3$-fold increase in scale leads to a new set of problems. % that lead towards our specific contribution in the current work. 
+Sequence mappers based on minimizer chaining have mapping costs of $O(n \log n)$ for long sequences \citep{Jain_2022}, where $n$ is the number of seeds.
+The vast number of minimizers and minimizer chains found when mapping whole chromosomes into a pangenome causes a dramatic increase in overall runtime, particularly when mapping repetitive sequences that are now accessible using improved 3rd-generation sequencing technology \citep{Logsdon_2021,Nurk_2022}.
+
+Here we show that, just as with previous leaps in assembly resolution, we can radically improve alignment performance for this new scale by further the degree to which we sparsify, or sample, key information driving mapping and alignment.
+We present wfmash, an ultra-scalable sequence aligner for whole genomes and pangenomes.
+%Our goals are different: we seek to build alignments that represent synteny across whole chromosomes, and we are less concerned about 
+To obtain approximate mappings between pairs of genomes, wfmash extends MashMap \citep{Jain_2018}, which has proven to be the fastest homomoly mapping method, adapting it to use world minimizers, and extending it with a new mapping chaining approach designed to find maximal syntenic blocks even in the presence of large structural variants.
+However, obtaining a base level alignment required a completely new approach to sequence alignment.
+We exploit the $O(s^2)$ scaling properties of the WaveFront Algorithm (WFA) \citep{Marco_Sola_2020} in a new hierarchical arrangement (WF$\lambda$) wherein WFA is applied not to the matching of single characters of input genomes, but rather to the matching of segments of hundreds of base-pairs (e.g.
+256bp).
+These segments are matched by sketching their kmers, and where similarity is above a given threshold, an ``incepted'' WFA alignment is completed at the base level.
+By taking half-overlapping segments, the method has a high probability of finding the optimal base-level alignment within the incepted WFA problems, and we ultimately combine these during traceback through the WF$\lambda$ matrix.
+A final round of erosion and patching of the resulting alignment (with WFA) yields a base-level alignment.
+For a segment length $L$, wavefront inception provides $O(4L^2)$ reduction in alignment complexity relative to direct application of WFA.
+When $L=256$, this corresponds to a 16384-fold reduction in computational complexity, a property born out by our experimental results.
+
+%A collection of minimizers of a given length is often referred to as a sketch, and is a form of locality sensitive hash that may be quickly compared to obtain sequence similarity estimates without consideration of the full sequence, as in mash \cite{Ondov_2016}.
+%Minimizers have been applied to numerous problems in sequence analysis, but the 
+%
 

### 2022-04-01 manuscript/main.tex
tweaks in intro

@@
 %Our goals are different: we seek to build alignments that represent synteny across whole chromosomes, and we are less concerned about 
-To obtain approximate mappings between pairs of genomes, wfmash extends MashMap \citep{Jain_2018}, which has proven to be the fastest homomoly mapping method, adapting it to use world minimizers, and extending it with a new mapping chaining approach designed to find maximal syntenic blocks even in the presence of large structural variants.
+To obtain approximate mappings between pairs of genomes, wfmash extends MashMap \citep{Jain_2018}, which has proven to be the fastest homomoly mapping method, adapting it to use world minimizers, extending it with a new mapping chaining approach designed to find maximal syntenic blocks even in the presence of large structural variants.
+
 However, obtaining a base level alignment required a completely new approach to sequence alignment.
-We exploit the $O(s^2)$ scaling properties of the WaveFront Algorithm (WFA) \citep{Marco_Sola_2020} in a new hierarchical arrangement (WF$\lambda$) wherein WFA is applied not to the matching of single characters of input genomes, but rather to the matching of segments of hundreds of base-pairs (e.g.
-256bp).
+We exploit the $O(s^2)$ scaling properties of the WaveFront Algorithm (WFA) \citep{Marco_Sola_2020} in a new hierarchical arrangement (the ``we flying'' alignment algorithm, WFlign) wherein WFA is applied not to the matching of single characters of input genomes, but rather to the matching of segments of hundreds of base-pairs (e.g.
+256bp) (WF$\lambda$).
 These segments are matched by sketching their kmers, and where similarity is above a given threshold, an ``incepted'' WFA alignment is completed at the base level.
@@
 For a segment length $L$, wavefront inception provides $O(4L^2)$ reduction in alignment complexity relative to direct application of WFA.
-When $L=256$, this corresponds to a 16384-fold reduction in computational complexity, a property born out by our experimental results.
+When $L=256$, this corresponds to a 16,384-fold reduction in computational complexity, a property born out by our experimental results.
+Although WFlign is an approximate alignment algorithm not guaranteed to find the optimal alignment through the full matrix, here we show that our implementation nevertheless achieves state-of-the-art accuracy for variant detection via alignment.
 

### 2022-04-01 manuscript/main.tex
typo

@@
 $^{\text{\sf 3}}$Department of Computer Sciences, Barcelona Supercomputing Center, Barcelona, 08034, Spain \\
-$^{\text{\sf 4}}$Departament d’Arquitectura de Computadors i Sistemes Operatius, Universitat Autònoma de Barcelona, Barcelona, 08193, Spain \\
+$^{\text{\sf 4}}$Department d’Arquitectura de Computadors i Sistemes Operatius, Universitat Autònoma de Barcelona, Barcelona, 08193, Spain \\
 $^{\text{\sf 5}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA

### 2022-04-01 manuscript/main.tex
typos

@@
 However, increasing sequence read lengths, and the arrival of whole genome assemblies for many species rendered such approaches more sensitive than necessary to achieve near-perfect alignment performance when mapping long reads to reference genomes.
-Methods like minimap2 \citep{Li_2018}, LRA \citep{Ren_2021}, Winnowmap \citep{Jain_2020}, and MashMap \citep{Jain_2018} compute mappings, or candidate alignments, using a sparse subset of all $k$-mers, which are often referred to as ``minimizers'' due to their definition as hash-minimizing $k$-mers in a given window \cite{Roberts_2004}.
+Methods like minimap2 \citep{Li_2018}, LRA \citep{Ren_2021}, Winnowmap \citep{Jain_2020}, and MashMap \citep{Jain_2018} compute mappings, or candidate alignments, using a sparse subset of all $k$-mers, which are often referred to as ``minimizers'' due to their definition as hash-minimizing $k$-mers in a given window \citep{Roberts_2004}.
 Rather than considering all $k$-mers, \textit{windowed minimizer} sampling collects $\sim L/w$ $k-mers$ from a sequence of length $L$, wherein each $k$-mer minimizes a given hash function within a window of size $w$.
 This sampling scheme provides guarantees about the distribution of sampled $k$-mers, which aids read alignment and mapping chaining.
-However, it has been discovered that window minimziers are biased and may underestimate the true similarity of sequences \citep{Belbasi_2022}.
+However, it has been discovered that window minimizers are biased and may underestimate the true similarity of sequences \citep{Belbasi_2022}.
 In contrast, \textit{world minimizers}, which sparsify over the entire potential space of $k$-mers \citep{Broder_1997}, are not biased, and have proved useful as the basis for the whole genome comparison system Mash \citep{Ondov_2016}.

### 2022-04-02 manuscript/fig_wflamba.tex
introduction
tweaks

@@
-\begin{figure}[ht!]
+\begin{figure}[t!]
     \begin{subfigure}{\linewidth}
@@
     \caption{
-      \textit{wflign} exploration of the alignment of $\sim$560kbp homologus segments of two \textit{Escherichia coli} genomes (GenBank NC\_000913.3 and NC\_008563.1).
+      WFlign exploration of the alignment of $\sim$560kbp homologus segments of two \textit{Escherichia coli} genomes (GenBank NC\_000913.3 and NC\_008563.1).
       Gray indicates cells explored by WF$\lambda$.

### 2022-04-02 manuscript/main.tex
tweaks

--- 2022-04-01 manuscript/main.tex
@@
 
-\title[Whole-genome pairwise alignment]{Whole-genome pairwise alignment using the hierarchical wavefront algorithm}
+\title[Wavefront inception]{Whole genome alignment with wavefront inception}
 \author[Guarracino \textit{et~al}.]{
@@
 
-However, increasing sequence read lengths, and the arrival of whole genome assemblies for many species rendered such approaches more sensitive than necessary to achieve near-perfect alignment performance when mapping long reads to reference genomes.
+Increasing sequence read lengths and the arrival of whole genome assemblies for many species rendered such approaches more sensitive than necessary.
+%The new problem of mapping long reads to reference genomes led to 
 Methods like minimap2 \citep{Li_2018}, LRA \citep{Ren_2021}, Winnowmap \citep{Jain_2020}, and MashMap \citep{Jain_2018} compute mappings, or candidate alignments, using a sparse subset of all $k$-mers, which are often referred to as ``minimizers'' due to their definition as hash-minimizing $k$-mers in a given window \cite{Roberts_2004}.
@@
 This sampling scheme provides guarantees about the distribution of sampled $k$-mers, which aids read alignment and mapping chaining.
-However, it has been discovered that window minimziers are biased and may underestimate the true similarity of sequences \citep{Belbasi_2022}.
+It has been discovered that window minimziers are biased and may underestimate the true similarity of sequences \citep{Belbasi_2022}.
 In contrast, \textit{world minimizers}, which sparsify over the entire potential space of $k$-mers \citep{Broder_1997}, are not biased, and have proved useful as the basis for the whole genome comparison system Mash \citep{Ondov_2016}.
@@
 The vast number of minimizers and minimizer chains found when mapping whole chromosomes into a pangenome causes a dramatic increase in overall runtime, particularly when mapping repetitive sequences that are now accessible using improved 3rd-generation sequencing technology \citep{Logsdon_2021,Nurk_2022}.
+
+%Minimizer seeds have also proven useful for the generation of alignments.
+%To obtain base level alignments using adaptive-banded Smith-Waterman-Gotoh requires seeds that initiate the correct alignment, rendering the use of minimizers necessary.
+%But, this limits the level of further sparsification available to mapping methods.
+%These costs are further exacerbated when obtaining the base level alignments.
+
+\subsection{Current work}
 
@@
 A final round of erosion and patching of the resulting alignment (with WFA) yields a base-level alignment.
+
 For a segment length $L$, wavefront inception provides $O(4L^2)$ reduction in alignment complexity relative to direct application of WFA.
@@
 %
+
+\input{fig_wflamba}
 
@@
 Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments \textit{W}-bp long from the two sequences to align.
-
-\input{fig_wflamba}
 

### 2022-04-02 manuscript/main.tex

@@
 Increasing sequence read lengths and the arrival of whole genome assemblies for many species rendered such approaches more sensitive than necessary.
+Sparsification, or random sampling to reduce computational costs, allows tremendous improvement in performance.
+We can still achieve near-perfect alignment performance when mapping long reads to reference genomes even when ignoring many parts of the query and target sequences.
 %The new problem of mapping long reads to reference genomes led to 

### 2022-04-03 manuscript/main.tex
tweaks

@@
 \textbf{Motivation:}
-Pairwise alignment is essential to understanding DNA sequence variation.
-The time and memory required to compute pairwise alignments increases quadratically with sequence length and nucleotide divergence.
+Homology mapping and base-level alignment are essential to understanding DNA sequence variation.
+The time and memory required to compute pairwise alignments increases quadratically with sequence length and nucleotide divergence, necessitating specialized heuristic approaches.
 Advances in sequencing technology have resulted in the rapid production of new whole-genome sequences.
@@
 Our approach achieves a many thousand-fold reduction in time and space complexity relative to direct alignment with WFA.
-We demonstrate that this comes at little to no cost in terms of variant detection accuracy relative to established sequence aligners. \\
+We demonstrate that this comes at little to no cost in terms of variant detection accuracy relative to established sequence aligners, and that our approach supports efficient pangenome-scale alignment.
+\\
 %\textit{wfmash} allows us to scale to all-to-all alignment of hundreds of vertebrate \\
@@
 Methods like minimap2 \citep{Li_2018}, LRA \citep{Ren_2021}, Winnowmap \citep{Jain_2020}, and MashMap \citep{Jain_2018} compute mappings, or candidate alignments, using a sparse subset of all $k$-mers, which are often referred to as ``minimizers'' due to their definition as hash-minimizing $k$-mers in a given window \cite{Roberts_2004}.
-Rather than considering all $k$-mers, \textit{windowed minimizer} sampling collects $\sim L/w$ $k-mers$ from a sequence of length $L$, wherein each $k$-mer minimizes a given hash function within a window of size $w$.
+Rather than considering all $k$-mers, \textit{windowed minimizer} sampling collects $\sim N/w$ $k$-mers from a sequence of length $N$, wherein each $k$-mer minimizes a given hash function within a window of size $w$.
 This sampling scheme provides guarantees about the distribution of sampled $k$-mers, which aids read alignment and mapping chaining.
@@
 Although WFlign is an approximate alignment algorithm not guaranteed to find the optimal alignment through the full matrix, here we show that our implementation nevertheless achieves state-of-the-art accuracy for variant detection via alignment.
+We exploit this dramatic reduction in computational complexity to develop a method capable of the direct pairwise alignment of chromosome-scale homologies.
+This arrangement yields 1:1 alignments that are critical for the development of pangenome graph models, multiple whole genome alignments, and projection---or lift-over---of annotations between assemblies.
 

### 2022-04-24 manuscript/main.tex
remove bioinformatics references
nsf ppos funding

@@
 
-\subtitle{Genome Analysis}
+\subtitle{ } %Genome Analysis}
 
@@
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+\history{ } %Received on XXXXX; revised on XXXXX; accepted on XXXXX}
 
-\editor{Associate Editor: XXXXXXX}
+\editor{ } %Associate Editor: XXXXXXX}
 
@@
 
-We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG) and efforts by Dr. Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy (AG).
+We gratefully acknowledge support from NIH/NIDA U01DA047638 and NSF PPoSS Award \#2118709 (EG) and efforts by Dr.
+Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy (AG).
 \linebreak
